### 2014-08-29 main_text.tex
Initial commit

\documentclass[11pt, letterpaper]{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}

% Math, graphics, and bibliography
\usepackage{amsmath,amssymb,amsfonts,mathrsfs,mathtools}
\usepackage{graphicx}
\usepackage[round,numbers,sort&compress]{natbib}
\renewcommand{\bibnumfmt}[1]{#1.}

% My preferred fonts
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[sc]{mathpazo}

% For fancier fractions and figure captions
\usepackage[labelfont={bf}, margin=1cm]{caption}
\usepackage{units}

% \usepackage[utf8]{inputenc}
% \usepackage[english]{babel}
% \usepackage{booktabs}

% \numberwithin{equation}{section}

\usepackage{authblk}

\begin{document}
\title{Amplitude metrics for cellular \\ circadian bioluminescence reporters}
\author[1]{Peter C. St. John}
\author[2]{Stephanie R. Taylor}
\author[1]{John H. Abel}
\author[1,*]{Francis J. Doyle III}
\affil[1]{Department of Chemical Engineering, University of California Santa
Barbara, Santa Barbara, California 93106-5080}
\affil[2]{Department of Computer Science, Colby College, Waterville, Maine
04901}
\affil[*]{Email: \texttt{doyle@engineering.ucsb.edu}}
% \author{}
\date{\today}
\maketitle

\begin{center}
Running head:\\ {Amplitude response curves for circadian systems} \\[1ex]
Keywords:\\ Systems Biology | Ordinary Differential Equations \\ Gene
Regulatory Network | Stochastic | Synchronization
\end{center}

\pagebreak
% 300 words
\begin{abstract}
Bioluminescence rhythms from cellular reporters have become the most common method used to quantify oscillations in circadian gene expression.
These experimental systems can reveal phase and amplitude change resulting from circadian disturbances, and can be used in conjunction with mathematical models to lend further insight into the mechanistic basis of clock amplitude regulation.
However, bioluminescence experiments track the mean output from thousands of noisy, uncoupled oscillators, obscuring the direct effect of a given stimulus on the genetic regulatory network.
In many cases, it is unclear whether changes in amplitude are due to individual changes in gene expression level or due to a change in coherence of the population.
While such systems can be modeled using explicit stochastic simulations, these models are computationally cumbersome and limit analytical insight into the mechanisms of amplitude change.
We therefore develop theoretical and computational tools to approximate the mean expression level in large populations of non-interacting oscillators, and further define computationally efficient amplitude response calculations to describe phase-dependent amplitude change.
At the single-cell level, a mechanistic nonlinear ordinary differential equation (ODE) model is used to calculate the transient response of each cell to a perturbation, while population-level dynamics are captured by coupling this detailed model to a phase density function.
Our analysis reveals that amplitude changes mediated at either the individual-cell or population level can be distinguished in tissue-level bioluminescence data without the need for single-cell measurements.
We demonstrate the effectiveness of the method by modeling experimental bioluminescence profiles of light-sensitive fibroblasts, reconciling the conclusions of two seemingly contradictory studies.
This modeling framework allows a direct comparison between {\itshape in vitro} bioluminescence experiments and {\itshape in silico} ODE models, and will lead to a more quantitative understanding of the factors that affect clock amplitude.
\end{abstract}

\section*{Introduction}
In mammals, circadian rhythms are endogenous oscillations in gene transcription responsible for coordinating daily changes in physiology.
While the suprachiasmatic nucleus (SCN) in the brain serves as the body's master pacemaker, cells found in peripheral tissues also oscillate in a circadian manner \cite{Lamia2008}.
These peripheral clocks process systemic and SCN-mediated entraining cues, buffering against rhythmic changes in energy availability to maintain metabolic homeostasis \cite{Kornmann2007}.
The amplitude of circadian transcription is a relevant factor, and has been shown to play a critical role in phase resetting and entrainment \cite{Pittendrigh1991, Abraham2010}.
Recent studies have further highlighted the importance of high peripheral clock amplitudes in maintaining metabolic health.
Mice lacking an intact clock have been shown to develop metabolic disease \cite{Marcheva2010}, while low amplitude clock oscillations, whether caused by diet \cite{Hatori2012} or age \cite{Chang2013} have also been tied to metabolic disorders.
An understanding of how clock amplitudes are regulated is therefore a topic of ongoing research with potential therapeutic applications \cite{St.John2014}.
 Unlike the network of oscillators in the SCN, in which intercellular coupling maintains robust amplitudes even in the absence of external cues, peripheral oscillators are thought to lack a direct mechanism to spontaneously synchronize \cite{Welsh2004}.
As a result, populations of peripheral clocks are likely synchronized by common external cues, with stochastic effects and cell heterogeneity driving entrained populations to gradually desynchronize.

The development of immortalized peripheral oscillator cell lines with bioluminescent reporters has allowed high-throughput analysis of the responses of circadian rhythms to genetic and pharmacological manipulation \cite{Hirota2010, Ramanathan2014}.
In addition to providing experimental tractability, these systems provide detailed information on the amplitude of oscillations in gene transcription, a measure often lacking from earlier experiments using wheel-running activity.
 As a result, these cell lines have proven useful in studying core clock connectivity and stoichiometry \cite{Baggs2009}.
However, since {\itshape in vitro} experiments typically measure entire cultures of cells, data collected at the population-level can obscure the response of the clock at the scale of the gene-regulatory network.
Even when individual cells can be recorded, stochastic noise hinders accurate amplitude determination.
As a result, two studies using similar perturbations to understand the mechanism of light-induced amplitude reduction reached differing conclusions, in which either single-cell amplitude reduction or population-level desynchrony was identified as the dominant factor \cite{Pulivarthy2007, Ukai2007}.

Mathematical models have long been used to understand the results of circadian experiments \cite{Leloup2003, Becker-Weimann2004}, aided by definitions and computational techniques designed to match modeling predictions to experimental data.
One such definition is the response function, a general technique that maps a change in an output variable to a temporary change in parameters \cite{Rand2004}.
For instance, the phase response curve (PRC) has been used to characterize the entrainment behavior of both experimental and mathematical systems \cite{Daan1976, Taylor2008, Pfeuty2011}, and in analyzing the synchrony of populations of oscillators \cite{Kim2014}.
Accurate and efficient numerical routines for finding infinitesimal PRCs have therefore been developed \cite{Kramer1984, Gunawan2006, Taylor2008a}.
In addition to changes in phase, phase-dependent changes in amplitude are an important factor in understanding circadian rhythms.
Amplitude response curves (ARCs) were first used in the clock literature to understand the effects of light pulses on simple phase-amplitude models, and were useful in predicting phase singularity behavior \cite{Jewett1998}.
ARCs have also been used to characterize perturbations to groups of oscillators through desynchrony \cite{Achermann1999, Ukai2007}, at the level of the single oscillator \cite{VanderVeen2012, Castejon2013}, and even in studying the effect of entrainment phase on SCN rhythm amplitude \cite{VanOosterhout2012}.
However, previous definitions of the ARC are inconsistent between studies and do not simultaneously consider amplitude effects at the single-cell and population level.

The choice of modeling framework dictates the type of amplitude response which can be predicted.
Ordinary differential equation (ODE) models of gene regulation are capable of describing the amplitude and phase-resetting behavior of single cells, but fail to capture the collective dynamics of a population oscillators.
Likewise, phase-only models correctly capture the change in synchrony of a population, but do not capture fluctuations in amplitude of individual oscillators.
Explicit stochastic simulations of populations of cells are capable of realistically capturing both single-cell and population-level effects, and have been successfully used to understand the response of coupled oscillators to external VIP perturbation \cite{An2013}.
However, these methods are computationally expensive, and cannot be used for an analytical understanding of amplitude response.

In this study, we describe approaches to quantify amplitude change in a population of non-interacting oscillators.
By exploiting the independence of each oscillator, we derive computationally efficient methods to approximate the mean dynamics of full stochastic simulations.
Additionally, our method allows the calculation of ARCs at both the single-cell and population level, allowing the behavior of the system to be quickly profiled.
Specifically, we use ODE models to describe the transient amplitude response at the single-cell level, coupled with a phase probability density function to describe population-level dynamics.
Following a perturbation with a finite duration, a limit cycle oscillator undergoes a transient change in amplitude.
When the mean expression level of a population of oscillators is also considered, an additional change in amplitude is incurred due to the change in synchrony of the population, which persists until synchrony is changed by subsequent perturbations.
We therefore observe a separation in timescales between the effects on clock output mediated at the single-cell level and those mediated by population synchrony, allowing the source of an amplitude change to be qualitatively inferred by inspection of bioluminescence data.
Understanding the mechanisms and consequences of both types of amplitude regulation will be important in understanding how peripheral amplitudes are maintained, and may lead to the design of pharmacological or behavioral strategies to boost circadian amplitudes.

\section*{Materials and Methods}

\subsection*{Basic definitions}

Ordinary differential equation (ODE) models take the general form
\begin{equation}
  \frac{dx}{dt} = f(x(t), p)
  \label{eq:odefn}
\end{equation}
in which $x(t)$ represents the concentrations of the state variables, such as mRNA and protein concentrations, $f$ contains information on the production, degradation, and reactivity of the states, and $p$ are the kinetic parameters which govern reaction kinetics.
Limit cycle models are ODE models in which the solution approaches a steady state oscillatory trajectory, satisfying the equation: \begin{equation}
  \lim_{t \to \infty} \left[ x(t + T) - x(t) \right] = 0
  \label{eq:limit}
\end{equation}
The period is the smallest $T > 0$ for which Eq.~\ref{eq:limit} holds.
The points on the stable limit cycle are denoted by $x^\gamma(\theta)$ with each point assigned to a value of a phase variable $\theta \in [0, 2\pi)$.
For convenience, time in Eq.~\ref{eq:odefn} can be rescaled such that the period is $2\pi$:
\begin{equation}
  \tilde{t} = \frac{2\pi}{T}t; \quad \tilde{f} = \frac{T}{2\pi}\;f; \quad \frac{dx}{d\tilde{t}} = \tilde{f}(x(\tilde{t}), p)
  \label{eq:that}
\end{equation}
The phase variable $\theta$ is therefore defined on the limit cycle as $\theta = \tilde{t}\mod 2\pi$, with $\theta = 0$ assigned to a unique and identifiable point.

\subsection*{Perturbations to limit cycle systems}

In this study we restrict our analysis to temporary perturbations capable of entraining an oscillatory system, excluding permanent parameter changes arising, for instance, from genetic knockout experiments.
The most simple entraining perturbation involves adding or removing components to a limit cycle system, resulting in a perturbed trajectory $x(t)$.
Here the initial conditions are determined by the strength of the perturbation and the phase at which it is applied:
\begin{equation}
  x(0) \coloneqq x^\gamma(\theta_0) + \Delta x(0)
  \label{eq:stateperturbation}
\end{equation}
This trajectory evolves according to Eq.~\ref{eq:odefn}, eventually returning to $x^\gamma$.
It is useful to express this trajectory by the deviation from the limit cycle:
\begin{equation}
  \Delta x(\tilde{t}) = x(\tilde{t}) - x^\gamma(\tilde{t} + \theta_0)
  \label{eq:delxt}
\end{equation}

In addition to perturbations directly to the state of the system, oscillators can also be perturbed by temporary changes to the parameters.
For a parameter pulse $\Delta p$ of duration $\tilde{d}$ (in radians) that ends at $\theta_0$, the oscillator deviates from the limit cycle trajectory according to:
\begin{align}
  x_{\tilde{d}}(-\tilde{d}) &= x^\gamma(\theta_0 - \tilde{d}) \\
  \frac{dx_{\tilde{d}}}{d\tilde{t}} &= \tilde{f}(x_{\tilde{d}}(\tilde{t}), p + \Delta p) \label{eq:ode_pert}
\end{align}
The pulse trajectory $x_{\tilde{d}}(\tilde{t})$ is then integrated from $\tilde{t} = -\tilde{d} \to 0$, at which point the pulse is removed.
A temporary parameter pulse is therefore equivalent to a state perturbation, but with the perturbation at $t = 0$ defined by:
\begin{equation}
  \Delta x(0) = x_{\tilde{d}}(0) - x^\gamma(\theta_0)
\end{equation}
A schematic depicting an example perturbed and reference trajectory is shown in Fig. 1A.

The response of the system to very small perturbations is more efficiently calculated using ODE sensitivity analysis \cite{Rabitz1983}.
The sensitivity matrix:
\begin{equation}
  S(t) = \frac{dx(t)}{dx(0)} = \lim_{\Delta x(0) \to 0}\frac{\Delta x(t)}{\Delta x(0)}
  \label{eq:senslimit}
\end{equation}
can be calculated directly by integrating:
\begin{equation}
  \frac{d}{dt} S(t)  = \frac{df(x(t),p)}{dx}\; S(t)
  \label{eq:odesens}
\end{equation}
with $S(0) = I$.

\subsection*{Phase response curves}
Quantifying phase changes following a perturbation has been well studied and is particularly relevant for models of circadian rhythms \cite{Kramer1984, Taylor2008a}.
A phase response curve (PRC) maps the change in phase resulting from the same perturbation applied at each initial phase.
Infinitesimal PRCs, the derivative of the phase change with respect to the perturbation, can be defined for state and parameter-impulse perturbations \cite{Taylor2008a}.
Methods for efficiently calculating these quantities using ODE sensitivity analysis have been developed \cite{Taylor2008a}, with the important result that the parameter- and state-impulse PRCs can be related by the Jacobian matrix:
\begin{equation}
  \frac{d}{d\tilde{t}}\frac{d\theta}{dp} = \frac{d\theta}{dx}\frac{d\tilde{f}}{dp} 
  \label{eq:pPRCequiv}
\end{equation}
This result follows from the fact that in the limit of an infinitely short and small parameter pulse, $\Delta x(0) \to \nicefrac{d\tilde{f}}{dp}\,\tilde{d}\, \Delta p$.

\subsection*{Phase-diffusion model}
Large populations of oscillators are typically described using phase-only models \cite{Rougemont2006}, in which the state of each oscillator is represented only by its phase, $\theta$.
The synchrony of the population can be modeled using a probability density function $p(\theta, \tilde{t})$ that describes the probability of finding an oscillator at each phase \cite{Kuramoto1984}.
The usefulness of probability density functions in describing the phase and amplitude responses of populations of circadian cells has been previously shown \cite{Ukai2007}.
As with all probability density functions:
\begin{equation}
  \int_0^{2\pi} p(\theta, \tilde{t}) \; d\theta = 1
\end{equation}
The shape of $p(\theta, \tilde{t})$ changes as the cells advance in time.
Stochastic effects cause the population to gradually desynchronize as slight cycle-to-cycle deviations are propagated throughout the population.
For a infinite population of oscillators, these effects are well-described by a Fokker-Plank equation \cite{Stein1965}:
\begin{equation}
  \frac{\partial p}{\partial \tilde{t}} = \frac{\partial p}{\partial \theta} + d\frac{\partial^2 p}{\partial \theta^2}\\
  \label{eq:pde}
\end{equation}
Due to the rescaling of $t$, the mean period of the population is $2\pi$.
Here, the $\nicefrac{\partial p}{\partial \theta}$ term, analogous to convection, describes the mean oscillatory period, while the $\nicefrac{\partial^2 p}{\partial \theta^2}$ term describes the diffusion of phases across $[0, 2\pi)$.
The phase diffusivity parameter $d$ (in units of inverse radians) describes the speed with which the population desynchronizes and can be fit to experimental data \cite{Rougemont2007}.
Eq.~\ref{eq:pde} has periodic boundary conditions, with initial condition $\phi(\theta)$ as the phase population at $t=0$:
\begin{align}
  \text{BCs:}\quad p(0, \tilde{t}) &= p(2\pi, \tilde{t}) \\
  \frac{\partial p}{\partial \theta}(0, \tilde{t}) &= \frac{\partial p}{\partial \theta}(2\pi, \tilde{t}) \\
  \text{IC:}\quad p(\theta, 0) &= \phi(\theta)\label{eq:pde_ic}
\end{align}
The solution of Eqs.~\ref{eq:pde}-\ref{eq:pde_ic} is well-characterized, with $p(\theta, \tilde{t})$ evolving in time as the convolution of the initial conditions with a wrapped normal distribution with mean $\tilde{t}$ and standard deviation $\sqrt{2d\tilde{t}}$ \cite{Chirikjian2009}:
\begin{equation}
  p(\theta, \tilde{t}) = \phi(\theta) * \mathcal{WN}(\theta; \tilde{t},
  \sqrt{2\, d\, \tilde{t}})
\end{equation}
in which the wrapped normal distribution \cite{Mardia2009} is defined as:
\begin{equation}
  {\cal WN}(\theta; \mu, \sigma) =
  \frac{1}{\sigma\sqrt{2\pi}}\sum_{k=-\infty}^\infty \exp\left[\frac{-(\theta
  - \mu + 2\pi k)^2}{2\sigma^2}\right]
  \label{eq:wrapped_normal}
\end{equation}
Since the convolution of two normal distributions is also a normal distribution, it is efficient when possible to describe $\phi(\theta)$ as a normal distribution with mean $\mu_0$ and standard deviation $\sigma_0$, such that $p(\theta, \tilde{t})$ can be found analytically through:
\begin{equation}
  p(\theta, \tilde{t}) = \mathcal{WN}(\theta; \mu_0 + \tilde{t},
  \sqrt{\sigma_0^2 + 2d^2\tilde{t}^2})
\end{equation}

\subsection*{Numerical simulations}

ODE models were simulated in Python, using the computer algebra package CasADi \cite{Andersson2013b}.
Stochastic simulations were performed using the StochKit2 package \cite{Sanft2011a}.
The codes used to generate the figures are included as a supplemental file.

\section*{Results and Discussion}

A signal to a population of limit cycle oscillators can affect amplitude in two ways.
First, individual cells are perturbed from their limit cycles, and exhibit transient dynamics before settling back to steady state amplitudes.
Secondly, the phases of the population are changed, resulting in a permanent change in population synchrony.
Therefore, in order to derive continuous approximations to the dynamics of a large population of non-interacting oscillators, we first demonstrate how ARCs can be found at both the single-cell and population level.

\subsection*{Definition of an amplitude metric at the single-cell level}

Following a perturbation to a single cell, the path that the perturbed trajectory $x(t)$ takes in returning to the limit cycle will have a different amplitude than the unperturbed trajectory.
While such an amplitude change will only have a finite duration, it plays an important role when perturbations are repeatedly received by the clock, such as a peripheral oscillator entrained to daily metabolic stimuli.
To define an amplitude change metric for such a case, we compare a perturbed trajectory $x(t)$ to a phase-shifted limit cycle reference $y(t)$, for which $x(t) \to y(t)$ for sufficiently long times:
\begin{equation}
  \begin{aligned}
    \Delta A (x(t), y(t)) &\coloneqq \int_0^\infty (x(t) - \mu)^2 - (y(t) - \mu)^2 \; dt\\
    &= \int_0^\infty h(t) \; dt
  \end{aligned}
  \label{eq:ampchangedefinition}
\end{equation}
This amplitude metric was chosen over alternatives, such as peak-trough distance, for its analytical tractability.
Since $x(t)$ approaches the reference as $t \to \infty$, the means of both trajectories are equal.
The mean of the limit cycle can therefore be calculated as:
\begin{equation}
  \mu \coloneqq \int_0^{2\pi} \frac{x^\gamma(\theta)}{2\pi} \; d\theta
  \label{eq:mu}
\end{equation}
The integrand in Eq.~\ref{eq:ampchangedefinition}, abbreviated by $h(t)$, compares the variances of the reference and perturbed trajectories.
When $h(t) > 0$, the trajectory is further from the mean than the reference, and similarly when $h(t) < 0$ the trajectory is closer to the mean than the reference.
Thus the overall amplitude change can be calculated by integrating $h(t)$ until the two trajectories converge, returning an amplitude value for each state variable.
Amplitude change for a two-dimensional oscillator is easy to visualize graphically.
In Fig.~1 B-D, the same state perturbation $\Delta x(0)$ applied at two different phases results in opposite amplitude changes, depending on whether the perturbation shifts the trajectory to the interior of the limit cycle (reduced amplitudes) or to the outside the limit cycle (increased amplitudes).
Trajectories for the second state variable, $Y(\tilde{t})$, and the corresponding integrand for the amplitude change equation, $h(\tilde{t})$, demonstrate how this transient change is quantified.

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[width=.75\textwidth]{figures/figure_1.pdf}
  \end{center}
  \caption{
{\bfseries Amplitude metrics at the single-cell level measure transient deviations from the limit cycle} ({\bfseries A}) Schematic showing trajectories used in the calculation of single-cell phase and amplitude change.
A perturbation $x(\tilde{t},p)$, blue, from the limit cycle solution $x^\gamma(\tilde{t},p)$, red, ultimately returns to the limit cycle with a phase shift $x^\gamma(\tilde{t} + \Delta\theta,p)$, green.
({\bfseries B-D}) The same perturbation applied at two different phases can result in opposite amplitude effects.
The state-space representation ({\bfseries B}) reveals the path both perturbations take to return to the limit cycle.
The time-series representation ({\bfseries C}) shows how the perturbation in blue results in an amplitude decrease, while the one in red results in an amplitude increase.
({\bfseries D}) This amplitude change is quantified by integrating $h(\tilde{t})$, the difference in variance from each solution to the limit cycle as defined in Eq.~\ref{eq:ampchangedefinition}, from $t=0 \to \infty$.
Here the shaded region indicates the area under the curve, $\Delta A$, for the perturbations shown in red and blue.
Model adapted from \cite{Novak2008}.
} \end{figure}

\subsection*{Single-cell amplitude response curves}
Similar to the PRC, we denote the phase-dependent amplitude change following a perturbation as an amplitude response curve (ARC), using the metric presented in Eq.~\ref{eq:ampchangedefinition}.
In Fig.~2, we calculate the PRC and ARC for a state perturbation of three different strengths.
The weakest perturbation results in type 1 (weak) resetting, in which the phase response curve is continuous, while the strongest perturbation results in type 0 (strong) resetting, in which the PRC is discontinuous.
This transition occurs once the state perturbation is strong enough to push the trajectory over the unstable fixed point located at the middle of the limit cycle.
For a perturbation of intermediate strength, at a critical phase the trajectory will be pushed close to the fixed point and take a long time to recover the steady state amplitude.
The behavior at this singularity point is indicated by a sharp dip in the ARC for this perturbation strength.

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[width=.75\textwidth]{figures/figure_2.pdf}
    \caption{
{\bfseries Phase and amplitude response curves describe how a single oscillator will respond to a perturbation.}  ({\bfseries A-B}) The phase and amplitude response curves for three perturbations of increasing strengths demonstrate the transition from type 1 phase resetting for a weak stimulus (blue) to type 0 phase resetting from a strong stimulus (red).
({\bfseries C}) The state space representation for the perturbations of varying strengths demonstrate how the trajectories return to the limit cycle.
For the particular perturbation shown in green, the trajectory starts very close to the singularity, and therefore takes a long time to recover.
({\bfseries D}) For an oscillator perturbed to the singularity, a strong amplitude reduction ensues.
Here, the oscillator with an initial phase $\theta \approx \nicefrac{3\pi}{2}$ is perturbed by $\Delta x = 1.1$ at $\tilde{t} = 0$.
Several cycles are required for normal amplitudes to be restored, corresponding to a dip in the ARC (C, green).}
  \end{center}
\end{figure}

Infinitesimal versions of response curves are often more general and easier to compute than those that track a specific perturbation strength \cite{Rand2004}.
We therefore derive an expression for the infinitesimal ARC, defined as:
\begin{equation}
  \frac{dA}{dx} \coloneqq \lim_{\Delta x(0) \to 0} \frac{\Delta A\left(x(\tilde{t}),\ x^\gamma(\tilde{t} + \theta_0 + \Delta\theta)\right)}{\Delta x(0)} = \int_0^\infty \lim_{\Delta x(0) \to 0} \frac{h(t)}{\Delta x(0)} \; dt
  \label{eq:sARC}
\end{equation}
While this quantity could be calculating by using a very small $\Delta x(0)$, it is more accurate and efficient to derive a direct method to calculate Eq.~\ref{eq:sARC} using the ODE sensitivities defined in Eq.~\ref{eq:odesens}.
For simplicity, we define $t_\theta = \tilde{t} + \theta_0$, which allows the time variable for the perturbation to vary from $0 \to \infty$ while tracking the appropriate phase on the limit cycle.
Since $\Delta \theta \to 0$ as $\Delta x(0) \to 0$, we Taylor expand the limit cycle trajectory around $t_\theta$:
\begin{align}
  x^\gamma(t_\theta + \Delta\theta) &= x^\gamma(t_\theta) +
  \frac{dx^\gamma(t_\theta)}{d\theta}\Delta\theta + O(\Delta\theta^2)\\
  &= x^\gamma(t_\theta) + \tilde{f}\left(x^\gamma(t_\theta)\right)\Delta\theta +
  O(\Delta\theta^2)
\end{align}
Simplifying the integrand in Eq.~\ref{eq:sARC}:
\begin{align}
  \frac{h(t)}{\Delta x(0)} &= \frac{1}{\Delta x(0)} \left[\left(x^\gamma(t_\theta) +
    \Delta x(\tilde{t}) - \mu\right)^2 - \left(x^\gamma(t_\theta +
    \Delta \theta) - \mu\right)^2 \right]\\
  &= \frac{1}{\Delta x(0)} \left[\left(x^\gamma(t_\theta) +
    \Delta x(\tilde{t}) - \mu\right)^2 - \left(x^\gamma(t_\theta) + \tilde{f}\left(x^\gamma(t_\theta)\right)\Delta\theta - \mu\right)^2 \right]\\
  &= \frac{1}{\Delta x(0)} \left[\left(\Delta x(\tilde{t}) -
    \tilde{f}\left(x^\gamma(t_\theta)\right)\Delta\theta\right) \left(\Delta
    x(\tilde{t}) + \tilde{f}\left(x^\gamma(t_\theta)\right)\Delta\theta +
    2(x^\gamma(t_\theta) - \mu)\right) \right]
\end{align}
Taking the limit of this integrand as $\Delta x(0) \to 0$ cancels several differential terms, and allows the remainder to be substituted with quantities that can be calculated using ODE sensitivity analysis:
\begin{align}
  \lim_{\Delta x(0) \to 0} \frac{h(t)}{\Delta x(0)} &= 2\left(\lim_{\Delta x(0) \to 0}\frac{\Delta x(\tilde{t})}{\Delta x(0)} - \tilde{f}\left(x^\gamma(t_\theta)\right)\lim_{\Delta x(0) \to 0}\frac{\Delta\theta}{\Delta x(0)}\right) \left(x^\gamma(t_\theta) - \mu\right)\\
  &= 2\left(S(\tilde{t}) - \tilde{f}(x^\gamma(t_\theta))\frac{d\theta}{dx}\right)\left(x^\gamma(t_\theta) - \mu\right)
  \label{eq:sens_sub}
\end{align}
Note that in Eq.~\ref{eq:sens_sub}, $\nicefrac{d\theta}{dx}$ represents the derivative of $\Delta\theta$ with respect to the perturbation, and is therefore a scalar quantity.
The infinitesimal state-impulse ARC may therefore be calculated directly from the ODE sensitivity matrix and the PRC, allowing the amplitude change for an infinitesimal perturbation to be calculated exactly and efficiently:
\begin{equation}
  \frac{dA}{dx} = \int_0^\infty 2\left(S(\tilde{t}) - \tilde{f}(x^\gamma(t_\theta))\frac{d\theta}{dx}\right)\left(x^\gamma(t_\theta) - \mu\right) \; dt
  \label{eq:siARC}
\end{equation}
Here, the first term of the integrand $(S - \tilde{f}\dot{\theta})$ tracks the distance from the perturbed trajectory to the limit cycle, which decays to zero as $t \to \infty$.
The second term $(x^\gamma - \mu)$ weights this distance by whether or not the deviation occurs above or below the oscillatory mean, yielding negative amplitude changes when the trajectory is perturbed closer to the mean.
Just as with the parameter-impulse PRC, the infinitesimal parameter-impulse ARC is defined as:
\begin{equation}
  \frac{d}{d\tilde{t}}\frac{dA}{dp} \coloneqq \lim_{\tilde{d},\; \Delta p \to 0}
  \frac{\Delta A}{\tilde{d}\, \Delta p}
\end{equation}
and may be calculated from the state-impulse version with the following relationship:
\begin{equation}
  \frac{d}{d\tilde{t}}\frac{dA}{dp} = \frac{dA}{dx}\frac{d\tilde{f}}{dp}
    \label{eq:piARC}
\end{equation}
As with Eq.~\ref{eq:pPRCequiv}, this equivalency reflects the fact that for a pulse of infinitely short duration, a parameter change is equivalent to changing the state of the system along the direction specified by the Jacobian.
Convergence between the methods in Eqs.~\ref{eq:siARC}-\ref{eq:piARC} and finite-difference approaches is shown in Fig.~S1 in the Supporting Material, which demonstrates that the numerically-efficient differential ARCs remain representative even for moderate-strength perturbations.

\subsection*{Population-level response curves}

We next briefly describe how to calculate the PRC and ARC for a phase-density model.
These approaches are commonly used in understanding phase models \cite{Kuramoto1984, Ukai2007}, and are presented here to match previous definitions for single cells.
A phase transition curve, $g(\theta) = \theta + \Delta\theta$, maps the phase of an oscillator before perturbation to its phase after the perturbation.
Since individual oscillators are neither created nor destroyed during the perturbation, it is possible - yet numerically difficult - to directly calculate the phase probability distribution following perturbation using the standard change of variables relation:
\begin{equation}
  \hat{p}(\theta, \tilde{t})\; dg(\theta) = p(\theta, \tilde{t})\; d\theta
  \label{eq:pdfinversion}
\end{equation}
However, it is easier to estimate the mean and standard deviation of the perturbed population $\hat{p}(\theta, \tilde{t})$ using directional statistics \cite{Mardia2009}.
A population defined on the unit circle can be described by a complex variable $z = \rho e^{i\bar{\theta}}$, where $\bar{\theta}$ is the mean phase and $\rho$, the synchronization index, is related to the standard deviation of the population.
For $\rho = 1$, the population is clustered about one mean phase, while for $\rho = 0$ the population is evenly balanced across the unit circle.
Complex variables for the population before and after perturbation can be calculated via:
\begin{align}
  z &\coloneqq \int_0^{2\pi} e^{i\theta} p(\theta, \tilde{t}) \; d\theta \label{eq:zbar}\\
  \hat{z} &\coloneqq  \int_0^{2\pi} e^{i\theta} \hat{p}(\theta, \tilde{t}) \; d\theta =
  \int_0^{2\pi} e^{i g(\theta)} p(\theta, \tilde{t}) \; d\theta
  \label{eq:zhat}
\end{align}
In Eq.~\ref{eq:zhat}, we avoid calculating the perturbed population explicitly by instead integrating over the new phases at the prior population density function.
Population-level amplitude and phase responses can therefore be calculated by:
\begin{align}
  \Delta \bar{\theta} &= \angle z - \angle \hat{z} \\
  \Delta \rho &= |z| - |\hat{z}|
  \label{eq:popampchange}
\end{align}
Population-level PRCs and ARCs can be tabulated by solving Eqs.~\ref{eq:zbar}-\ref{eq:popampchange} for populations $p(\theta, \tilde{t})$ with different mean phases.
It is important to note that the ARC at the population level strongly depends on the slope of the PRC, as has been shown previously \cite{Ukai2007}.

\subsection*{Population-level mean expression profiles}

To efficiently capture the population-level effects of bioluminescence experiments, we couple the detailed single-cell ODE model to a phase-density model.
Previous work has used limit cycle models to estimate population-level parameters, such as desynchronization rate, for phase-only models \cite{Rougemont2007}.
In this study, we use an ODE model to calculate the response to a perturbation at each phase, and subsequently take the weighted average of these responses according to the phases of the cells in the population.

Assuming each oscillator in the population follows the dynamics described by $x^\gamma(\theta)$, the unperturbed mean population-level expression, $\bar{x}(\tilde{t})$, can be found by taking the weighted average of the expression level over the current population:
\begin{equation}
  \bar{x}(\tilde{t}) = \int_0^{2\pi} x^\gamma(\theta) p(\theta, \tilde{t}) \;
  d\theta
  \label{eq:xbar}
\end{equation}

Phase-diffusion models can explain why the gradual damping from experimental population-level data closely resembles an exponentially damped sinusoid, a result that has been shown experimentally \cite{Welsh2004} and computationally \cite{Rougemont2007}.
Our method similarly demonstrates exponential decay: for the idealized system $x^\gamma(\theta) = \cos(\theta)$ starting from a synchronized state:
\begin{align}
  \bar{x}(\tilde{t}) &= \frac{1}{\sqrt{4\pi d\tilde{t}}}\int_{-\infty}^{\infty}
  \cos(x) \exp\left(-\frac{(x - \tilde{t})^2}{4d\tilde{t}}\right)\; dx \\
  &= \Re\left[\frac{1}{\sqrt{4\pi d\tilde{t}}}\int_{-\infty}^{\infty}
    \exp(ix) \exp\left(-\frac{(x - \tilde{t})^2}{4d\tilde{t}}\right)\; dx\right] \\
    &= \Re\left[e^{(i - d)\tilde{t}}\right]\\
    &= e^{-d\tilde{t}} \cos(\tilde{t}) \label{eq:expsin}
\end{align}
Due to the smoothing effect of phase diffusion, higher frequency sinusoidal components of the limit cycle are damped faster than lower frequency components, resulting in an exponentially decaying sinusoids even for limit cycles that are not sinusoidal in sufficiently disperse populations.
The analytical result in Eq.~\ref{eq:expsin} allows us to easily estimate the phase diffusivity parameter in Eq.~\ref{eq:pde} by simply fitting an exponentially damped sinusoid to detrended bioluminescence data.

Using our continuous approximation to population-level dynamics, we demonstrate an example unperturbed trajectory using Model 1 (adapted from \cite{Novak2008}, see Supporting Material).
In Fig.~3, an initially jagged population density smooths and widens over time as cells desynchronize, similar to the effects of diffusion.
While each cell's expression level follows the limit cycle, the population amplitude gradually damps with time as cells with diverse phases are averaged together.

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[width=.75\textwidth]{figures/figure_3.pdf}
    \caption{
{\bfseries Synchrony effects population-level amplitude.}
({\bfseries A}) Stochastic fluctuations cause a population of cells to gradually desynchronize with time.
The phase probability density, $p(\theta, \tilde{t})$, gradually widens as it advances in phase according to the mean period.
({\bfseries B}) The mean amplitude from a population of oscillators is determined by the probability density function $p(\theta, \tilde{t})$ and the oscillator's limit cycle $x^\gamma(\theta)$.
As time passes, population-level rhythms resemble an exponentially damped sinusoid.}
  \end{center}
\end{figure}

Next we describe how to calculate population-level mean expression following a perturbation.
The perturbed trajectory $\hat{x}(\tilde{t})$ can be decomposed into contributions from two sources.
First, for long times after the perturbation, each individual oscillator will have returned to the limit cycle $x^\gamma(\theta)$, but with a new population density $\hat{p}(\theta, \tilde{t})$.
This steady-state perturbed trajectory $\hat{x}_{ss}(\tilde{t})$ can be found by:
\begin{equation}
  \hat{x}_{ss}(\tilde{t}) = \int_0^{2\pi} x^\gamma(\theta) \hat{p}(\theta, \tilde{t}) \; d\theta
  \label{eq:xhatss}
\end{equation}
For very long times following perturbation, the new phase probability density could be approximated using the initial mean and standard deviation found through Eq.~\ref{eq:zhat}, as jagged profiles in the population density will eventually smooth to a normal distribution.
However, for shorter times following perturbation, $\hat{p}(\theta, \tilde{t})$ must be calculated numerically.

The second contribution to the perturbed population trajectory comes from deviations from limit cycle oscillations in each cell.
We calculate the population-level effect of these deviations by averaging over the deviations that occur at each phase.
We define the deviation trajectory $\delta x(\theta, \tilde{t})$ for each phase as the distance between the perturbed trajectory and the phase-adjusted reference:
\begin{align}
  \delta x(\theta_0, \tilde{t}) &\coloneqq x(\tilde{t}) - x^\gamma(\tilde{t} + \theta_0 + \Delta \theta) \\
  \therefore\; \lim_{\tilde{t} \to \infty} \delta x(\theta, \tilde{t}) &= 0
\end{align}
Since the perturbed trajectory ultimately converges with the phase-adjusted reference, deviations will converge to zero.
Since the phase change, $\Delta\theta$, associated with a perturbation at each phase is likely not known prior to calculating the perturbed trajectory, it is difficult to tabulate deviation trajectories associated with each final phase, $\theta_0 + \Delta\theta$.
It is therefore more straightforward to find the average effect of single-cell perturbations at the population level by weighting the deviations by the phase density function prior to perturbation.
The population-level response to a perturbation, $\hat{x}(\tilde{t})$, is therefore defined as:
\begin{equation}
  \hat{x}(\tilde{t}) = \int_0^{2\pi} x^\gamma(\theta)\hat{p}(\theta, \tilde{t}) + \delta x(\theta, \tilde{t})p(\theta, \tilde{t}) \; d\theta
\end{equation}
Here the first term is equivalent to $\hat{x}_{ss}(\tilde{t})$, the steady-state perturbed trajectory, while the second term decays to zero as individual oscillators return to their steady-state amplitude.
The accuracy of the continuous phase-diffusion model was tested by explicitly simulating a perturbation to a population of 225 uncoupled oscillators using a stochastic simulation algorithm \cite{Gillespie1977, Sanft2011a}.
Good agreement between the stochastic population and continuous approximation was shown in phase shift, single-cell level amplitude change, and alteration of population synchrony (Fig.~S2, Movies S1 \& S2).
These results indicate the use of a continuous probability function is justified in the case of cultured cellular reporter systems, which may contain up to $10^5$ individual cells per culture \cite{Welsh2004}.
In addition, the method was verified by similar simulations using the Oregonator \cite{Field1974}, a model containing only mass-action terms, and a detailed mechanistic model of circadian rhythms \cite{Hirota2012} at a variety of phase timings (Fig.~S3).
The continuous approximation is therefore able to capture the population-level dynamics of a wide variety of limit cycle models and perturbations.

We next demonstrate the usefulness of single-cell and population level ARCs in predicting the mean response of a population.
Using a detailed single-cell model (Model 2), we calculate the response of the system, specifically the average CRY2 protein expression level, to a temporary increase in {\itshape Per} transcription rate.
In Fig.~4A, we plot the resulting single-cell and population-level response curves.
In this case, the population-level phase change is a slightly smoothed version of the single-cell PRC, since the population has an averaging effect on incoming perturbations with each cell receiving the input at a slightly different internal phase.
This smoothing follows from the consistent definition of phase between the single-cell and population level.
In the ARC plot, however, the shape of the single-cell and population-based ARC are different, since they describe different types of changes (finite vs.\ sustained) in the output trajectories.

Using these response curves, we demonstrate characteristic features of amplitude change mediated at the single-cell and population-level.
In Fig.~4B the perturbation at $\theta_1$ does not cause a significant change in phase or population synchrony, yet strongly reduces single-cell amplitudes.
The resulting population-level trajectory appears damped for the first cycle, until the oscillators within the population return to the limit cycle and normal amplitudes are restored.
In Fig.~4C the perturbation at $\theta_2$ reduces population synchrony while increasing single-cell amplitudes, yielding amplitudes that are transiently increased before the population settles to a lower-amplitude, desynchronized trajectory.

These examples demonstrate the qualitative differences between amplitude change mediated at the single-cell and population level.
Their characteristic features allow us to distinguish between these two sources of amplitude change without explicitly recording single-cell amplitudes: amplitude change induced at the population-level will be sustained, but may be masked in the short term by changes to single-cell level amplitudes.
Similarly, a change induced at the single-cell level will be evident only at short time-scales.
These results underscore the importance in considering both single-cell and population-mediated amplitude change in predicting the effects of daily stimuli on clock amplitudes.

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[width=.75\textwidth]{figures/figure_4.pdf}
    \caption{
{\bfseries Response curves describe different types of amplitude change.}
A population of cells is simulated using a detailed model of circadian rhythms (Model 2).
Each cell in the population is subjected to a $20\%$ increase in the $\mathit{vtp}$ parameter for $d=\nicefrac{\pi}{2}$, inducing both a phase and amplitude change.
({\bfseries A}) The phase response curves are shown for both the single limit cycle oscillator and population average, left.
The population-level ARC is shown together with the single-cell amplitude response, with both curves normalized to $\sigma=1$.
({\bfseries B-C}) Population-level rhythms, $\hat{x}(\tilde{t})$ resulting from perturbations given at the indicated phases in A (red and cyan lines, respectively).
The population transitions from the unperturbed population, $\bar{x}(\tilde{t})$, to the steady-state perturbed population, $\hat{x}_{ss}(\tilde{t})$.
({\bfseries B}) A perturbation at this phase yields a transient amplitude reduction resulting from perturbations at the single-cell level, however, there is little change in population synchrony.
Population-level rhythms are therefore transiently damped before regaining normal amplitudes.
({\bfseries C}) Oscillators are desynchronized, but with a transient increase in limit cycle amplitude.
It therefore takes some time before the amplitude reduction from desynchrony is realized.}
  \end{center}
\end{figure}

\subsection*{Application to Experimental Data}

We conclude by using the proposed modeling framework to reconcile two experimental studies that measured the effect of light on the circadian amplitude of photosensitive fibroblast cells \cite{Ukai2007, Pulivarthy2007}.
Both studies sought to determine the main factor of amplitude reduction following a transient light pulse, in which either desynchrony alone \cite{Ukai2007} or a combination of single-cell amplitude reduction and desynchrony \cite{Pulivarthy2007} was identified as the dominant factor.

Data on the response of melanopsin-responsive and control NIH3T3 mouse fibroblasts cells is shown in Fig.~5B \cite{Ukai2007}.
The experimental data is de-noised using a discrete wavelet transform \cite{Leise2011}.
In the experiment, a light pulse is given to desynchronize a colony of cells, resulting in drastically reduced amplitudes.
A second (longer) light pulse subsequently re-synchronizes the population, resulting in increased amplitudes.
To capture this phenomenon {\itshape in silico}, we used a recent model of the core circadian feedback circuit \cite{Hirota2012}.
We assumed an initial phase population with $\sigma = 0.1$ and calculated a phase diffusivity parameter $d = 0.104$ from the experimental control trajectory.

The next step in capturing the experimental data was finding a perturbation capable of desynchronizing the system.
To find such a perturbation, we calculated population-level ARCs at several different knockdown strengths for each parameter in the model.
We selected the degradation rate of {\itshape Per} mRNA from the feasible parameters, due to PER's known induction by CREB following photo-perturbation \cite{Tischkau2003}.
The parameter was reduced by $28.5\%$ during the light pulses, a value fit to maximize light-induced desynchrony.
The output variable, PER2-luciferase in the experimental system, was chosen to be represented by {\itshape Cry1} mRNA - an E box activated gene that is buffered from the direct effects of the parameter perturbation.
The single-cell and population-level response curves for this parameter choice are shown in Fig.~5A, which demonstrates the strong desynchronization that occurs at $\theta\approx\nicefrac{\pi}{4}$.
Since the reporter used in the experimental system likely has a phase lag from the corresponding mRNA, we did not try to match the initial phase of the simulation to experiment.
Instead, the initial phase of simulation was chosen such that the first pulse occurred when the system was at the phase that corresponded with the minimum of the population-level ARC.

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[width=.75\textwidth]{figures/figure_5.pdf}
    \caption{
{\bfseries Method allows direct comparison between model results and experimental bioluminescence profiles.} 
({\bfseries A}) Response curves for the parameter perturbation used to model the first light pulse in (B).
The initial phase of the system is chosen to coincide with the strong dip in the population-level ARC (left).
({\bfseries B}) De-noised bioluminescence data from Ukai {\itshape et al.} 2007 (top) demonstrates population-level amplitude change resulting from two light pulses \cite{Ukai2007}.
The first pulse desynchronizes the system, while the second resynchronizes.
These perturbations were reproduced {\itshape in silico} (bottom) using the model from Hirota {\itshape et al.} 2012.
The experimental results are qualitatively captured by the model, demonstrating the suitability of the method in capturing bioluminescence experiments.
The relative amplitude of the model equations has been scaled to match that of the normalized bioluminescence profiles.
({\bfseries C}) Example probability density functions for the model trajectory shown in (A).
 The initial population (red) is effectively desynchronized (orange) following the first light pulse.
Following the second light pulse, stronger peaks are seen (teal), which gradually damp as time evolves.} \end{center}
\end{figure}

The simulated control and perturbed trajectories, Fig.~5B, closely match the experimental results, in which the model captures both the phase shifts and amplitude modulation of the light pulses.
The phase probability density function for the light-sensitive model trajectory is shown in Fig.~5C for several representative time points.
Changes to the synchronization of the population from each perturbation are readily apparent, with both perturbations inducing a bimodal distribution in the phase density.
While experimental data on individual cell phases was unavailable for this data set, bimodal distributions in SCN neuron firing following a phase shift have previously been seen experimentally \cite{Rohling2011}.
As higher-frequency features, these bimodal distributions dissipate as the phases diffuse.

Several quantitative differences between the model and experiment do appear.
Most importantly, the acute amplitude change following the light pulse is not correctly captured by the single-cell model.
From the experimental system, it appears the light pulse temporarily reduces PER2-luciferase amplitudes immediately following perturbation, separate from the overall change in population synchrony.
This result is most obvious after the second light pulse, where rhythms require some time before their maximum amplitude is reached.
Such a delay is indicative of perturbations to the limit cycle system, suggesting a contribution from single-cells in determining the population amplitude.
Light-induced amplitude suppression in fibroblasts is therefore likely mediated at the single-cell level at short time-scales, and at the population level for longer time-scales.
To improve the fit of the model to the data, the response curves could be fit to match experimentally predicted values.
Infinitesimal PRCs and ARCs calculated at the single-cell level are numerically efficient to evaluate, and could be more readily incorporated into a parameter estimation algorithm than explicit stochastic simulations of a population of oscillators.

\section*{Conclusion}

In this study, we have described new tools for understanding amplitude change in independent cells based on analytically tractable methods for simulating a large population of oscillators.
These tools synthesize single-cell level ODE sensitivity analysis metrics with population-level measures of synchrony, allowing the quantification of clock amplitude at multiple levels of biological organization.
Using these tools, we have demonstrated how ODE models can be directly compared to experimental bioluminescence profiles, which currently serve as a standard system to investigate clock perturbations.
More generally, we have demonstrated the differences between acute and prolonged changes in amplitude following temporary perturbations, and characterized the mechanisms that give rise to each.

While our examples have focused on light-mediated perturbations, these approaches could also be applied to pharmacological perturbations.
Some difficulty arises in obtaining such data in cultured cells, however, as entraining perturbations would require that pharmacological agents be introduced for only a finite duration.
As medium or temperature changes are often enough to resynchronize cultured cells, a transient application is difficult to achieve experimentally.
The search for clock-enhancing molecules has therefore tended to focus on constant drug concentrations: for instance, dose-dependent period or amplitude change following inhibition of CK1$\delta$ or similar targets \cite{Chen2013}.
For {\itshape in vivo} systems, however, pharmacokinetics dictates a finite duration of action for both naturally secreted hormones and pharmacological therapies.
More effective treatments might therefore be designed by explicitly accounting for such a transient response, perturbing peripheral clocks at the right phase to induce resynchronization and an increase in single-cell level amplitudes.

Maintaining robust rhythmicity in peripheral oscillators has been linked to protection against metabolic disease.
It is therefore interesting that liver cells have not developed a stronger mechanism of intercellular coupling, such as in the SCN, to maintain robust amplitudes in the absence of external cues.
Perhaps the ability to quickly re-entrain to a phase shift, which is often slowed by coupling \cite{Abraham2010}, has historically been more advantageous than protection against highly variable food intake.
Regardless, the necessities of modern society often dictate irregular circadian-metabolic regimes that damp peripheral clock oscillations.
For such instances, we hope that the mathematical theory presented here will permit the development of optimized treatment regimes.

\section*{Acknowledgments}
We thank Yongqiang Wang for the helpful discussions.
This work was supported by the National Institutes of Health/National Institute of General Medical Sciences under award number 1R01GM096873-01 and by the Institute for Collaborative Biotechnologies through grant W911NF-09-0001 from the U.S.
Army Research Office.

\bibliographystyle{biophysj.bst}
\bibliography{condensed_library.bib}

\end{document}

### 2014-08-29 supplemental_info.tex
Initial commit

\documentclass[11pt, letterpaper]{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}

% Math, graphics, and bibliography
\usepackage{amsmath,amssymb,amsfonts,mathrsfs}
\usepackage{graphicx}
\usepackage[round,numbers,sort&compress]{natbib}
\renewcommand{\bibnumfmt}[1]{#1.}
% \usepackage[square,sort,comma,numbers]{natbib} % my preferred formatting

% My preferred fonts
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[sc]{mathpazo}

% For fancier fractions and figure captions
\usepackage[labelfont={bf}, margin=1cm]{caption}
\usepackage{units}

% For model float
\usepackage{float}
\newfloat{model}{thp}{lop}
\floatname{model}{Model}
\floatstyle{plaintop}
\restylefloat{model}
\usepackage{supertabular}
\usepackage{booktabs}

\begin{document}

\begin{model}
\caption{Adapted from Nov\'{a}k \& Tyson, 2008 \cite{Novak2008}. Model used for
ARC demonstrations in Figs. 1-3, S1 ($P=4$) and Fig. S2, Movies S1 \& S2
($P=2$).}
  \centering
\begin{align*}
  \frac{dX}{dt} &= \frac{1}{1 + Y} - X \\
  \frac{dY}{dt} &= k_t \; X - k_d \; Y - \frac{Y}{\alpha_0 + \alpha_1 \; Y +
  \alpha_2 Y^2}
\end{align*}

\tablehead{\toprule Parameter & Value\\\midrule}
\begin{supertabular}{rl}
$k_t$ & 20 \\
$k_d$ & 1  \\
$P$   & 4 (or 2)  \\
\bottomrule\end{supertabular}\hspace{3ex}
\begin{supertabular}{rl}
$\alpha_0$ & 0.005 \\
$\alpha_1$ & 0.05  \\
$\alpha_2$ & 0.1   \\
\bottomrule\end{supertabular}
\end{model}

\begin{model}
\caption{Hirota {\itshape et al.} 2012 \cite{Hirota2012}. A more detailed model
of circadian rhythms used for simulations in Figs. 4-5.}
  \centering
\begin{align*}
\frac{d\,\mathit{p}}{dt} &= \frac{- \mathit{p} \cdot \mathit{vdp}}{\mathit{kdp} + \mathit{p}} + \frac{\mathit{vtp}}{\mathit{knp} + \left(\mathit{C1n} + \mathit{C2n}\right)^{3}}\\
\frac{d\,\mathit{c1}}{dt} &= \frac{- \mathit{c1} \cdot \mathit{vdc1}}{\mathit{c1} + \mathit{kdc1}} + \frac{\mathit{vtc1}}{\mathit{knc1} + \left(\mathit{C1n} + \mathit{C2n}\right)^{3}}\\
\frac{d\,\mathit{c2}}{dt} &= \frac{- \mathit{c2} \cdot \mathit{vdc2}}{\mathit{c2} + \mathit{kdc1}} + \frac{\mathit{vtc2}}{\mathit{knc1} + \left(\mathit{C1n} + \mathit{C2n}\right)^{3}}\\
\frac{d\,\mathit{P}}{dt} &= - \mathit{C1} \cdot \mathit{P} \cdot \mathit{vaC1P} + \mathit{C1n} \cdot \mathit{vdC1P} - \mathit{C2} \cdot \mathit{P} \cdot \mathit{vaC1P} + \mathit{C2n} \cdot \mathit{vdC1P} - \frac{\mathit{P} \cdot \mathit{vdP}}{\mathit{P} + \mathit{kdP}} + \mathit{ktxnp} \cdot \mathit{p}\\
\frac{d\,\mathit{C1}}{dt} &= - \mathit{C1} \cdot \mathit{P} \cdot \mathit{vaC1P} - \frac{\mathit{C1} \cdot \mathit{vdC1}}{\mathit{C1} + \mathit{kdC1}} + \mathit{C1n} \cdot \mathit{vdC1P} + \mathit{c1}\\
\frac{d\,\mathit{C2}}{dt} &= - \mathit{C2} \cdot \mathit{P} \cdot \mathit{vaC1P} - \frac{\mathit{C2} \cdot \mathit{vdC2}}{\mathit{C2} + \mathit{kdC1}} + \mathit{C2n} \cdot \mathit{vdC1P} + \mathit{c2}\\
\frac{d\,\mathit{C1n}}{dt} &= \mathit{C1} \cdot \mathit{P} \cdot \mathit{vaC1P} - \mathit{C1n} \cdot \mathit{vdC1P} - \frac{\mathit{C1n} \cdot \mathit{vdCn}}{\mathit{C1n} + \mathit{C2n} + \mathit{kdCn}}\\
\frac{d\,\mathit{C2n}}{dt} &= \mathit{C2} \cdot \mathit{P} \cdot \mathit{vaC1P} - \frac{\mathit{C2n} \cdot \mathit{MC2n} \cdot \mathit{vdCn}}{\mathit{C1n} + \mathit{C2n} + \mathit{kdCn}} - \mathit{C2n} \cdot \mathit{vdC1P}\\
\end{align*}

\tablehead{\toprule Parameter & Value\\\midrule}

\begin{supertabular}{rl}
$vtp $  & 0.195   \\
$vtc1$  & 0.131   \\
$vtc2$  & 0.114   \\
$knp $  & 0.425   \\
$knc1$  & 0.259   \\
$vdp $  & 0.326   \\
$vdc1$  & 0.676   \\
\bottomrule\end{supertabular}\hfil
\begin{supertabular}{rl}
$vdc2$ & 0.608   \\
$kdp $ & 0.0115  \\
$kdc1$ & 1.15    \\
$vdP $ & 2.97    \\
$vdC1$ & 0.0338  \\
$vdC2$ & 1.52    \\
$vdCn$ & 1.69    \\
\bottomrule\end{supertabular}\hfil
\begin{supertabular}{rl}
$MC2n $ & 2.02    \\
$kdP  $ & 0.101   \\
$kdC1 $ & 3.32    \\
$kdCn $ & 0.0526  \\
$vaC1P$ & 0.0406  \\
$vdC1P$ & 0.00175 \\
$ktxnp$ & 3.0     \\
\bottomrule\end{supertabular}
\end{model}

\begin{model}
  \caption{The Oregonator model \cite{Field1974}, with parameters as presented
  in \cite{Gillespie1977}. Used in Fig.~S3 as an example of a limit cycle
oscillator using only mass-action kinetics.}

  The individual chemical reactions (used for stochastic simulation) are:
\begin{align*}
  X_1 + Y_2 &\xrightarrow{c_1} Y_1\\
  Y_1 + Y_2 &\xrightarrow{c_2} Z_1\\
  X_2 + Y_1 &\xrightarrow{c_3} 2Y_1 + Y_3\\
  2Y_1 &\xrightarrow{c_4} Z_2\\
  X_3 + Y_3 &\xrightarrow{c_5} Y_2
\end{align*}

These reactions can be converted to a standard ODE model, in which only the
intermediate variables $Y$ are considered, as:
\begin{align*}
  \frac{d\,\mathit{Y1}}{dt} &= \mathit{c1x1}\cdot\mathit{Y2} -
  \mathit{c2}\cdot\mathit{Y1}\cdot\mathit{Y2} + \mathit{c3x2}\cdot\mathit{Y1} -
  \mathit{c4}\cdot\mathit{Y1}^2\\
  \frac{d\,\mathit{Y2}}{dt} &= -\mathit{c1x1}\cdot\mathit{Y2} -
  \mathit{c2}\cdot\mathit{Y1}\cdot\mathit{Y2} + \mathit{c5x3}\cdot\mathit{Y3}\\
  \frac{d\,\mathit{Y3}}{dt} &= \mathit{c3x2}\cdot\mathit{Y1} -
  \mathit{c5x3}\cdot\mathit{Y3}
\end{align*}

\begin{center}
  \tablehead{\toprule Parameter & Value\\\midrule}
  \begin{supertabular}{rl}
    $\mathit{c1x1}$ & 2 \\
    $\mathit{c2}$ & 0.1 \\
    $\mathit{c3x2}$ & 104 \\
    $\mathit{c4}$ & 0.016 \\
    $\mathit{c5x3}$ & 26 \\
    \bottomrule\end{supertabular}
\end{center}
\end{model}

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=.75\textwidth]{figures/figure_S1.pdf}
    \caption{{\bfseries Convergence of finite-difference and differential
    methods.} ({\bfseries A}) Perturbations of decreasing strength to the state
    of the oscillator result in phase and amplitude response curves that match
    the differential limit. ({\bfseries B}) Temporary perturbations of
    decreasing strength to a kinetic parameter. The duration of each
    perturbation is fixed to 0.2 radians. Finite difference approximations
    closely match the differential method, which is offset in phase by 0.1
    radians to account for the nonzero pulse duration.}
\end{center}
\end{figure}

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=.75\textwidth]{figures/figure_S2.pdf}
    \caption{{\bfseries Approximation of an explicit stochastic population by
    continuous methods.} ({\bfseries A}) A population of 225 stochastic
    oscillators was simulated using the Gillespie stochastic simulation
    algorithm (SSA), see Movie S1 and S2. The mean protein
    expression, $Y$, is plotted as a function of time. A $50\%$ reduction in
    the protein translation rate to each of the oscillators is applied from
    $\hat{t} = -\nicefrac{\pi}{4} \to 0$, resulting in both single-cell and
    population-level amplitude change (red solid line). This population is
    approximated by the continuous methods described in this manuscript, in
    which the decay parameter $d=0.025$ is estimated to match the
    stochastic-induced desynchrony of the control population (black solid
    line). The initial standard deviation $\sigma_0 = 0.48$ is similarly
    matched to the stochastic population. The resulting predicted unperturbed
    and perturbed trajectories, $\bar{x}(\hat{t})$ and $\hat{x}(\hat{t})$
    respectively, closely match the stochastically modeled values.  ({\bfseries
    B}) Phase histograms for the stochastic population are shown at several
    phases, both before and after the desynchronizing perturbation.  Phase
    probability-density functions for the continuous approximation are also
    shown, with close agreement between stochastic and continuous simulations.
    This close approximation validates the use of ODE models and
  phase-diffusion populations in deriving amplitude and phase-response behavior
for networks of uncoupled cells.}
\end{center}
\end{figure}

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/figure_S3.pdf}
    \caption{{\bfseries Validation of continuous methods with alternate
    models.} ({\bfseries A}) The Oregonator model (Model 3) was tested as an
    example of a mass-action limit cycle oscillator. A population of 2000
    oscillators was perturbed at size different mean phases, $\mu$, by
    increasing the $\mathit{c2}$ parameter by $50\%$ for $d=\pi$ (highlighted
    region). Only the average $\mathit{Y2}$ variable is plotted. ({\bfseries
    B}) The model described in \cite{Hirota2012} (Model 2) was similarly tested
    at a variety of mean phases by reducing the $\mathit{vdp}$ parameter by
    $28.5\%$ and plotting the resulting $\mathit{c2}$ state variable (as in
    Fig.~5). Parameters for the continuous approximations were found by
    estimating the phase, initial standard deviation, period, and phase
    diffusivity of the control population. Good agreement is seen between the
  continuous and stochastic simulations, indicating the proposed method is
suitable for predicting the population-level responses for a variety of model
types.}
\end{center}
\end{figure}

\clearpage
{\bfseries Movie S1:} Stochastic simulation of a two-state limit
cycle oscillator. A population of 225 stochastic oscillators was simulated
using the Gillespie stochastic simulation algorithm (SSA). The deterministic
limit cycle, $x^\gamma(\hat{t})$, is shown in black (left), and the current
location of individual stochastic oscillators are shown by green dots. The
population average, plotted in both the state (left) and time (right) domains,
is shown in red. Stochastic noise drives the population to gradually
desynchronize, resulting in damped oscillations of the average protein level.
\\[2ex]

{\bfseries Movie S2:} Perturbation to a population of stochastic
oscillators. The population of oscillators is perturbed by a $50\%$ reduction
in the protein translation rate to each of the oscillators from $\hat{t} =
-\nicefrac{\pi}{4} \to 0$ (yellow frames), resulting in both single-cell and
population-level amplitude change. The oscillators quickly recover to the limit
cycle (single-cell level amplitude change), but the population is left
desynchronized (population-level amplitude change).

\renewcommand{\refname}{Supporting References}
\bibliographystyle{biophysj.bst}
\bibliography{condensed_library.bib}
\end{document}

### 2014-08-29 main_text.tex
Sending to stephanie 2014-08-29

@@
  Unlike the network of oscillators in the SCN, in which intercellular coupling maintains robust amplitudes even in the absence of external cues, peripheral oscillators are thought to lack a direct mechanism to spontaneously synchronize \cite{Welsh2004}.
-As a result, populations of peripheral clocks are likely synchronized by common external cues, with stochastic effects and cell heterogeneity driving entrained populations to gradually desynchronize.
+As a result, populations of peripheral clocks are likely synchronized by common external cues, with stochastic effects and cell heterogeneity driving entrained populations gradually toward desynchrony.
 
@@
     \caption{
-{\bfseries Synchrony effects population-level amplitude.}
+{\bfseries Synchrony affects population-level amplitude.}
 ({\bfseries A}) Stochastic fluctuations cause a population of cells to gradually desynchronize with time.
@@
 A population of cells is simulated using a detailed model of circadian rhythms (Model 2).
-Each cell in the population is subjected to a $20\%$ increase in the $\mathit{vtp}$ parameter for $d=\nicefrac{\pi}{2}$, inducing both a phase and amplitude change.
+Each cell in the population is subjected to a $20\%$ increase in the $\mathit{vtp}$ parameter ({\itshape Per} transcription rate) for $d=\nicefrac{\pi}{2}$, inducing both a phase and amplitude change.
 ({\bfseries A}) The phase response curves are shown for both the single limit cycle oscillator and population average, left.
@@
 ({\bfseries B}) De-noised bioluminescence data from Ukai {\itshape et al.} 2007 (top) demonstrates population-level amplitude change resulting from two light pulses \cite{Ukai2007}.
-The first pulse desynchronizes the system, while the second resynchronizes.
+Circadian oscillations are suppressed by the first light pulse and rescued by the second.
 These perturbations were reproduced {\itshape in silico} (bottom) using the model from Hirota {\itshape et al.} 2012.

### 2014-08-29 main_text.tex
stephanie's changes, better explanations of the numerical methods in fig. 5

@@
   \hat{x}(\tilde{t}) = \int_0^{2\pi} x^\gamma(\theta)\hat{p}(\theta, \tilde{t}) + \delta x(\theta, \tilde{t})p(\theta, \tilde{t}) \; d\theta
+  \label{eq:xhat}
 \end{equation}
@@
 To capture this phenomenon {\itshape in silico}, we used a recent model of the core circadian feedback circuit \cite{Hirota2012}.
-We assumed an initial phase population with $\sigma = 0.1$ and calculated a phase diffusivity parameter $d = 0.104$ from the experimental control trajectory.
+In order to find parameters for the phase density function, we assumed an
+initial phase population with $\sigma = 0.1$ and calculated a phase diffusivity
+parameter $d = 0.104$ from the experimental control trajectory.
 
@@
 
+Population-level trajectories were found by solving for the resulting limit cycle trajectories at every phase, and subsequently finding the weighted average of these trajectories according the phase density (as described in Eqs.~\ref{eq:xbar} and \ref{eq:xhat}).
 The simulated control and perturbed trajectories, Fig.~5B, closely match the experimental results, in which the model captures both the phase shifts and amplitude modulation of the light pulses.

### 2014-09-02 response_letter.tex
Adding Latex version of response letter - few changes

\documentclass[11pt, letterpaper]{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}

% Math, graphics, and bibliography
\usepackage{amsmath,amssymb,amsfonts,mathrsfs,mathtools}

% My preferred fonts
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[sc]{mathpazo}
\usepackage{textcomp}

\usepackage{color}
\usepackage{tcolorbox}
\tcbuselibrary{breakable}
\tcbuselibrary{skins}
\usepackage{units}

\usepackage[parfill]{parskip} % For block paragraphs
\definecolor{gray}{gray}{0.4}
\newenvironment{reviewer}{\itshape\color{gray}}{}

\tcbset{skin=enhanced}

\newenvironment{manuscript}[1]{\begin{center}\begin{tcolorbox}[colback=green!5!white,colframe=green!75!black,width=0.8\textwidth,title={#1},breakable,fonttitle=\bfseries]}{\end{tcolorbox}\end{center}}

\begin{document}
\today\\

{\itshape Francis J. Doyle III}\\
Dept.\ of Chemical Engineering\\
Univ.\ of California, Santa Barbara\\
Santa Barbara, CA 93106-5080\\

{\itshape Leah Edelstein-Keshet}\\
Editorial Board Member\\
Biophysical Journal\\

Thank you for your email on August 1st, 2014, inviting us to submit a revised version of our manuscript, ``Amplitude metrics for cellular circadian bioluminescence reporters'' (MS: 2014BIOPHYSJ304278) by Peter C. St. John, Stephanie R. Taylor, John H. Abel, and Francis J. Doyle~III.

We thank the reviewers for their detailed comments of the manuscript and suggestions for how to improve the work. We did our best to balance the suggestions of Reviewer \#1, who suggested trimming much of the theoretical foundations of the method, with the suggestions of Reviewer \#2, who recommended additional explanations and validations. Ultimately we believe that the changes have greatly strengthened our manuscript by demonstrating our method is generalizable to alternative models and by focusing the work on our novel contributions.

Please see our detailed responses to the reviewers below:

\section*{Reviewer \#1}
\begin{reviewer}
In this manuscript the authors study the amplitude of oscillators with a focus on circadian rhythms.
Understanding the effect of perturbations of populations of oscillators is important to understand the mechanisms of circadian rhythms.

The authors start by describing phase response curves PRCs, ARCs and phase density equations.
They also describe practical approaches to computing these.
Using such a computed perturbed distribution, the authors show how to calculate an ``average oscillation'' of the population.

My main concern is that I cannot tell what is novel in the manuscript.
Since references are scant in the main body of the text, this is somewhat obscure.
For example, phase response curves have been discussed in numerous places, as have amplitude response curves.
The phase density methods and an associated Fokker-Planck equation (Eq.~31) also date back at least to Kuramoto and Stratonovich, with a lot of work following that.
For instance, the ``synchronization index'', $\rho$, is the order parameter in Kuramoto's treatment, and has been studied in hundreds of papers.
I guess this is well known, but some references should be given.
Stochastic oscillators at the population level, with a focus on PRCs, have been studied in a series of papers by Tanaka, Goldobin, Ermentrout and others - see also the standard, but somewhat dated book by Pikovsky, Rosenblum and Kurths.

Pages 4-14 of the manuscript therefore contain a lot of known results.
Some of the results may be new - for instance, I have not seen Eq.~(26) before - but the majority of the results presented here have been standard fare in the voluminous literature on stochastic oscillators.
Is the novelty here that these can be used to analyze data? If so, this is not well explained.
\end{reviewer}

We acknowledge that many of the mathematical formulas presented in the ``Results'' section have been used by many other studies: our intention was to introduce concepts and notations to readers who might be less familiar with the large body of previous literature.
The reviewer's point is a good one, however, and we have endeavored to draw as clear a line as possible between new concepts and derivations and those which come from the literature.
To this end, we have moved as much of the introductory derivations as possible to the ``Materials and Methods'' section, introducing previous work on phase-diffusion models before proceeding with sensitivity-based methods for finding the ARC.
Additionally, we have included additional references for historical derivations of many of the results.

To address the point directly, we consider the following contributions of the manuscript as novel, and highlight these in the results section:

\begin{itemize}
  \item Our definition of transient amplitude change (at the single-cell level). This definition allows us to calculate and directly compare amplitude change using both finite-difference and ODE sensitivity methods.
  \item The subsequent derivation of the differential ARC, similar to derivations of the differential phase response curve.
  \item We derive a method to approximate a stochastic population of noninteracting oscillators using continuous methods, including accounting for deviations from limit cycle trajectories.
\end{itemize}

While we have left some introductory derivations in the Results section (notably, derivations of the PRC and ARC for phase-only models), these sections have been clearly labeled.

\begin{reviewer}
The interesting part of the manuscript are the application to experimental data.
However, this part of the manuscript is very brief.
It is also not clear to me what methods discussed in the first part of the manuscript were used here.
An ARC was computed, and the phase density is also found for the model.
However, details on the connection to the first part of the manuscript are sparse.
\end{reviewer}

We have followed the reviewer's suggestion and expanded the section of the manuscript that focuses on the application to experimental data.
Specifically, we detail our method of selecting the parameter knockdown as well as the initial phase through the use of response curves.
In doing so, we have expanded Fig.~5 to include response curves for the first parameter pulse, demonstrating the strong dip in the population-level ARC.

\begin{reviewer}
Ultimately, I am not sure what the point of the manuscript is. The authors state that

``\dots we provide new tools and definitions to quantify amplitude change in a population of oscillators, and determine what factors dominate amplitude change under certain conditions.''

Then I disagree - most of the tools here are not new.
Density methods for uncoupled populations of oscillators have been around for many years - at least since Stratonovich.
What is novel is potentially the treatment of the amplitude response curve.
If this is indeed the novel contribution, then this needs to be better explained.
\end{reviewer}

We acknowledge that our placement of the introductory material for the phase density methods likely gave the impression that our major results were already well-known.
We believe our revised manuscript gives a better emphasis to the novel aspects of the study.

\begin{reviewer}
I suggest rewriting the manuscript with an emphasis on the application.
It is still a bit unclear how the proposed method can be applied to real data, and what one would get out of it.
Therefore a focus on these ideas, rather than the review of results about PRCs, ARCs and phase densities, would significantly improve the manuscript.
\end{reviewer}

We have modified extensive portions of the manuscript in order to make the novel contributions of the manuscript more clear, and added additional explanations to the application of the experimental data.

\begin{reviewer}
Minor comments: 
- Equation 11. Could $\Theta(x_0)$ be defined as $\theta$ such that $lim_{t\rightarrow \infty} | x(t)-x^\gamma(\theta+t)|=0$ ?
\end{reviewer}

Yes, although for length constraints (and to remove excess review of results on PRCs) this equation has been removed.

\begin{reviewer}
- Equation 16: Could $\mu$ just be defined as $\int_0^{2\pi} y(t)dt/(2\pi)$ ?
\end{reviewer}

\begin{reviewer}
- Equation 27 is incomplete. 
\end{reviewer}

We have fixed this regrettable error, and thank the reviewer for their careful reading.

\begin{reviewer}
- In Figure 5A, the initial condition for the experiments is 1. Why did the authors use the initial condition $0.5$ in the simulations?
\end{reviewer}

The initial conditions of the experiments likely arises from a normalization of the bioluminescence data, and absolute values were not available to directly compare amplitudes.
Therefore the relative amplitudes of both systems had to be compared, with the amplitude estimated from the control trajectory.
The phase offset of the experimental and model systems was also not fit exactly - we added an explanation of the rationale behind this decision to the manuscript:

\begin{manuscript}{Page 16}
The single-cell and population-level response curves for this parameter choice are shown in Fig.~5A, which demonstrates the strong desynchronization that occurs at $\theta\approx\nicefrac{\pi}{4}$.
Since the reporter used in the experimental system likely has a phase lag from the corresponding mRNA, we did not try to match the initial phase of the simulation to experiment.
Instead, the initial phase of simulation was chosen such that the first pulse occurred when the system was at the phase that corresponded with the minimum of the population-level ARC.
\end{manuscript}

\section*{Reviewer \#2}

\begin{reviewer}
The manuscript titled ``Amplitude metrics for cellular circadian bioluminescence reporters'' investigates a very interesting problem: amplitude perturbations via changes in single-cell levels and/or population synchrony.
Furthermore, the manuscript suggests that amplitude changes mediated by single-cell level or population-level changes can be distinguished based on the analysis in the manuscript.
I totally agree that this is a very important problem in circadian clocks as well as many other biological oscillators.
However, unlike the statement in the abstract of the manuscript ``In particular, our analysis reveals that amplitude changes mediated by individual cells or population synchrony can be distinguished in tissue-level bioluminescence data without the need for single-cell measurements'', I am not convinced how their analysis provides such a tool.
Furthermore, I found some mistakes in their derivation of a major formula, which reduces the reliability of the manuscript.
Finally, I am also concerned about major simulation results because the Gillespie algorithm with non-elementary propensity functions is used, which can cause a large error.
In summary, I strongly agree that this manuscript considers an interesting and important problem, but due to the above reasons, I am not convinced with their results.
To resolve these issues I have the following suggestions:

Major concerns

1) During the derivation of the major formula (Eq 18- 28), I found some errors. Please, fix them.
\end{reviewer}

We thank the reviewer for their careful review of the mathematical derivations. The errors are largely typographical:

\begin{reviewer}
a. (21) and (22) are not equal. 
\end{reviewer}

This mistake has been corrected.
Eq.~22 (now 27) is obtained by factoring the difference of squares from the previous equation, but we mistakenly kept a squared term from the previous step.

\begin{reviewer}
b. (22) and (23) are not equal. (23) is an approximation of (22) when delta theta approaches to zero. 
\end{reviewer}

This is a fair point, as we dropped the infinitesimal elements before taking the limit.
The equality is valid in the differential limit as the phase shift for an infinitesimal perturbation will be identically 0, and therefore this error in notation does not affect the validity of the results.
However, including the delta theta terms explicitly made the equation too long for a single line.
Since the equation (with the delta theta terms included) was largely a repetition of the previous equation, we have removed it.

\begin{reviewer}
c. (23) and (24) are not equal.
\end{reviewer}

This typo was also fixed by removing the squared term in Eq.~24 (now 29), which was unfortunately propagated from the typo in Eq.
22.

\begin{reviewer}
d. (27) is not complete.
\end{reviewer}

This error has been fixed.

\begin{reviewer}
e. Please, provide an intuitive explanation for formulas (26) and (28) for the reader. 
\end{reviewer}

The following two explanatory remarks have been added following Eq.~26 and 28 (now 30 and 32):

\begin{manuscript}{Page 10}
Here, the first term of the integrand $(S - \tilde{f}\dot{\theta})$ tracks the distance from the perturbed trajectory to the limit cycle, which decays to zero as $t \to \infty$.
The second term $(x^\gamma - \mu)$ weights this distance by whether or not the deviation occurs above or below the oscillatory mean, yielding negative amplitude changes when the trajectory is perturbed closer to the mean.
Just as with the parameter-impulse PRC, the infinitesimal parameter-impulse ARC is defined as:
\begin{equation}
  \frac{d}{d\tilde{t}}\frac{dA}{dp} \coloneqq \lim_{\tilde{d},\; \Delta p \to 0}
  \frac{\Delta A}{\tilde{d}\, \Delta p}
  \tag{31}
\end{equation}
and may be calculated from the state-impulse version with the following relationship:
\begin{equation}
  \frac{d}{d\tilde{t}}\frac{dA}{dp} = \frac{dA}{dx}\frac{d\tilde{f}}{dp}
  \tag{32}
\end{equation}
As with Eq.~11, this equivalency reflects the fact that for a pulse of infinitely short duration, a parameter change is equivalent to changing the state of the system along the direction specified by the Jacobian.
\end{manuscript}

\begin{reviewer}
2) Whereas the manuscript describes that they explore ``weakly coupled oscillators'', I found that what the manuscript is really considering are ``independent or non-interacting oscillators''.
Weakly coupled oscillators have been used in a different context (e.g.\ Kuramoto oscillators).
Please, correct this.
\end{reviewer}

The terminology has been changed throughout the manuscript.
We thank the reviewer for this correction.

\begin{reviewer}
3) One of the most interesting results of this manuscript (Fig4) is showing that ARCs, but not PRCs are very different at the single-cell level as compared to the population level.
I can see that a single simulation (Fig.~4A) supports this, however, I cannot see that the analysis in the manuscript is sufficient to support this statement.
That is, please explain how the analysis of this manuscript (perhaps the formula for the ARC) leads to the conclusion that appears in Fig.~4A.
Furthermore, I wonder if the result in Fig.~4A is a specific example of a special case or a general property.
In particular, a recent study (DeWoskin et al (2014)) shows that PRCs can also be very different in single-cell and population levels.
To test the generality of the results (Fig.~4A), please test this argument, at least with more simple models such as Model 1 in the manuscript, a Goodwin oscillator based on a Hill function, or the Kim-Forger model based on protein sequestration.
\end{reviewer}

Our point in discussing the phase response curves of Fig.~4A is only that phase maintains a consistent definition between the population and individual cell level, while amplitude must be defined differently in each case.
We have added the following text to the manuscript to make this point more explicit:

\begin{manuscript}{Page 14}
In this case, the population-level phase change is a slightly smoothed version of the single-cell PRC, since the population has an averaging effect on incoming perturbations with each cell receiving the input at a slightly different internal phase.
This smoothing follows from the consistent definition of phase between the single-cell and population level.
In the ARC plot, however, the shape of the single-cell and population-based ARC are different, since they describe different types of changes (finite vs.\ sustained) in the output trajectories.
\end{manuscript}

The PRC result in Fig.~4A is not a special case, but other types of phase-resetting behavior in a population can occur.
For instance, an almost desynchronized population would undergo type 0 (strong) phase resetting, even if the PRC of each cell was type 1 (weak), since any change to the phase-density profile would create a new mean phase.
Ultimately, we feel that the literature on the collective phase-response of a population of oscillators (coupled or non-coupled, simulated either explicitly or with phase-density methods) is relatively mature, and that a providing a detailed analysis of the phase responses we obtain is beyond the scope of our major contribution (outlined in response to Reviewer 1).

\begin{reviewer}
4) The author suggests that ``amplitude change induced at the population level will follow a consistent decaying sinusoid, while a change induced at the single-cell level will deviate from this exponential decay''.
Is this a general statement or the conclusion of a specific simulation result (Fig.~4B and C)? If this is not a general statement, then I doubt whether this result can be used to distinguish two sources of amplitude change without explicit recording of single cell amplitudes.
If this is a general statement, please provide more details and intuitive explanations, as well as further simulation results with more types of models, or analysis.
\end{reviewer}

This is a general statement regarding the dynamics of populations and limit-cycle systems. Ultimately, non-interacting oscillators have no driving force to re-entrain, so changes to synchrony remain long after limit-cycle oscillations are restored.

In addition to adding simulation results with additional models (see below), we have changed the quoted text to be more explicit in this explanation:

\begin{manuscript}{Page 14}
These examples demonstrate the qualitative differences between amplitude change mediated at the single-cell and population level.
Their characteristic features allow us to distinguish between these two sources of amplitude change without explicitly recording single-cell amplitudes: amplitude change induced at the population-level will be sustained, but may be masked in the short term by changes to single-cell level amplitudes.
Similarly, a change induced at the single-cell level will be evident only at short time-scales.
These results underscore the importance in considering both single-cell and population-mediated amplitude change in predicting the effects of daily stimuli on clock amplitudes.
\end{manuscript}

\begin{reviewer}
5) To derive the major formula in Eq.~(50), various approximations and assumptions were made.
This formula is tested with a single simulation.
To confirm the accuracy of Eq (50), please, test this in other models as suggested above.
\end{reviewer}

(Answered below)

\begin{reviewer}
6) Stochastic simulations of the model with non-elementary reactions (e.g.
Hill-functions or Michaelis-Menten terms) are done with the Gillespie algorithm.
However, recent studies have shown the inaccuracy of this approach (Thomas et al.
BMC Syst Biol (2012) and Agarwal et al.\ J Chem Phy (2012)).
Please, discuss whether this inaccuracy in the stochastic simulations affects the simulation results in Fig.~5.
If possible, please try a model based on mass-action kinetics or a model based on total QSSA since Yao et al.\ Biophy J (2008) shows that the model based on total QSSA can provide accurate simulations with the Gillespie algorithm.
\end{reviewer}

Points 5 and 6 are linked concerns regarding the simulation and approximation of stochastic trajectories, so we have addressed them together.
The noise generated using the Gillespie algorithm for non-elementary propensity functions can indeed be different than the noise that would occur for the full system when expanded into elementary reactions.
That being said, the level of noise present in Fig.~5 was chosen by varying the volume parameter of the stochastic simulation so that desynchronization rates matched experimentally observed data.
We therefore do not believe that incorrect noise estimates from the Gillespie algorithm affects our approximations of the stochastic population.
Regardless, we tested the method with two additional models.
First, the Oregonator, is a simple mass-action oscillatory reaction system considered in Gillespie's 1977 paper.
We used a perturbation to the conversion rate of two of the intermediate species at 6 different phases to test the accuracy of the perturbed approximation, $x_hat$.
Second, we tested the method with a stochastic simulation of a more detailed circadian model.
For consistency, we used the same model and perturbation as was used in Fig.~5 (for the first light pulse).
The results of these simulations are included as Fig.~S3, which (similar to Fig.~S2) demonstrates good agreement between the stochastic models and continuous approximations.

\begin{reviewer}
Minor concerns:

1) In most case, "hat" mark is used to represent various perturbed variables. However, "hat" mark is also used to represent the rescaled time variable. This can lead to confusion in understanding the equations. Please, use different notation for scaled time.
\end{reviewer}

This is a good suggestion, and we have changed the notation for scaled time to $\tilde{t}$ throughout the manuscript. Additionally, we noticed that our notation for pulse duration conflicted with our definition of phase diffusivity, so pulse duration has been changed to $\tilde{d}$.

\begin{reviewer}
2) Please provide a little more detail for the derivation of (14).
\end{reviewer}

The derivation for Eq.~14 (now 11) is analogous to Eq.~28 (now 32), a lengthy process that was detailed in the appendix of Taylor et al, 2008 (PMID 19593456).
For space constraints, we have added only brief remarks on their derivation:

\begin{manuscript}{Page 5}
This result follows from the fact that in the limit of an infinitely short and small parameter pulse, $\Delta x(0) \to \nicefrac{d\tilde{f}}{dp}\,\tilde{d}\, \Delta p$.
\end{manuscript}

\begin{reviewer}
3) In Fig.~2A, a discontinuous PRC appears as continuous curve.
Please, do not link the discontinuous points, and redraw the PRCs.
\end{reviewer}

This plot has been corrected.

\begin{reviewer}
4) The PRC and ARC in Fig 4A are in response to which perturbation?
Please provide details on how Fig 4A is simulated.
\end{reviewer}

Additional details on the simulation of Fig. 4 have been added to both the main text and the figure legend:

\begin{manuscript}{Page 14}
We next demonstrate the usefulness of single-cell and population level ARCs in predicting the mean response of a population.
Using a detailed single-cell model (Model 2), we calculate the response of the system, specifically the average CRY2 protein expression level, to a temporary increase in {\itshape Per} transcription rate.
In Fig.~4A, we plot the resulting single-cell and population-level response curves.
\end{manuscript}

\begin{manuscript}{Page 15}
Figure 4: Response curves describe different types of amplitude change.
A population of cells is simulated using a detailed model of circadian rhythms (Model 2).
Each cell in the population is subjected to a $20\%$ increase in the $\mathit{vtp}$ parameter ({\itshape Per} transcription rate) for $d=\nicefrac{\pi}{2}$, inducing both a phase and amplitude change.
\end{manuscript}

\begin{reviewer}
5) The definition of Eq (48) is not clear.
Please provide a more detailed explanation.
\end{reviewer}

Eq~ 48 (now 44) tracks the distance between the perturbed trajectory and the limit cycle for each phase.
The weighted-average of this distance is then used to find the population-level expression following perturbation.
To make this point more clear, we have added additional explanatory text to the manuscript:

\begin{manuscript}{Pages 12-13}
The second contribution to the perturbed population trajectory comes from deviations from limit cycle oscillations in each cell.
We calculate the population-level effect of these deviations by averaging over the deviations that occur at each phase.
We define the deviation trajectory $\delta x(\theta, \tilde{t})$ for each phase as the distance between the perturbed trajectory and the phase-adjusted reference:
\begin{align}
  \delta x(\theta_0, \tilde{t}) &\coloneqq x(\tilde{t}) - x^\gamma(\tilde{t} + \theta_0 + \Delta \theta)\tag{48} \\
  \therefore\; \lim_{\tilde{t} \to \infty} \delta x(\theta, \tilde{t}) &= 0 \tag{49}
\end{align}
Since the perturbed trajectory ultimately converges with the phase-adjusted reference, deviations will converge to zero.
Since the phase change, $\Delta\theta$, associated with a perturbation at each phase is likely not known prior to calculating the perturbed trajectory, it is difficult to tabulate deviation trajectories associated with each final phase, $\theta_0 + \Delta\theta$.
It is therefore more straightforward to find the average effect of single-cell perturbations at the population level by weighting the deviations by the phase density function prior to perturbation.
\end{manuscript}

\begin{reviewer}
6) In a couple of places, the manuscript discusses pharmacological manipulation of circadian clocks.
Could you discuss this in more detail since the manuscript analyzes data coming from only light-induced perturbations (Fig.~5)? 
For instance, the manipulation of circadian rhythms with the CK1d/e inhibitor (PF-670462) has been studied widely.
Can you discuss how your method can be applied to the data in these studies?
\end{reviewer}

Indeed, while we would have preferred to analyze data coming from pharmacological perturbations, this type of data is difficult to obtain.
We have added an additional paragraph to the conclusion to address this limitation, and point out potential future applications of the method:

\begin{manuscript}{Page 18}
While our examples have focused on light-mediated perturbations, these approaches could also be applied to pharmacological perturbations.
Some difficulty arises in obtaining such data in cultured cells, however, as entraining perturbations would require that pharmacological agents be introduced for only a finite duration.
As medium or temperature changes are often enough to resynchronize cultured cells, a transient application is difficult to achieve experimentally.
The search for clock-enhancing molecules has therefore tended to focus on constant drug concentrations: for instance, dose-dependent period or amplitude change following inhibition of CK1$\delta$ or similar targets (50).
For {\itshape in vivo} systems, however, pharmacokinetics dictates a finite duration of action for both naturally secreted hormones and pharmacological therapies.
More effective treatments might therefore be designed by explicitly accounting for such a transient response, perturbing peripheral clocks at the right phase to induce resynchronization and an increase in single-cell level amplitudes.
\end{manuscript}

\section*{Reviewer \#3}

\begin{reviewer}
Many circadian clock studies use reporters and analyze ensembles of cells.
Frequently phase shifts are analyzed using Phase Response Curves.
The authors complement this approach by introducing Amplitude Response Curves.
Moreover, they discuss effects of ensemble averaging.
A clear mathematical description is provided which is useful to interpret experimental data.
Some related studied should discussed to embed the manuscript and to highlight the novelty.

\end{reviewer}

We thank the reviewer for their positive impression of the manuscript, as well as for their helpful suggestions of related studies to discuss. We have incorporated many of the suggested works, which we feel helps to place the manuscript within the larger field of circadian modeling.

\begin{reviewer}
Detailed comments:

Felix Naef studied stochastic models to discuss fibroblast rhythms.
He found as well purely exponential decay in case of damped oscillators and deviations for limit cycles.
The current study should be compared to these models.
\end{reviewer}

In the relevant studies from Felix Naef, the authors use a phase model in which Gaussian white noise causes the frequencies to drift in time.
The decay rate is estimated from stochastic simulations of detailed limit cycle models, and used to capture unperturbed bioluminescence data.
We have incorporated two relevant references: Rougement et al, 2006 and Rougement et al, 2007, and added the references to the following sections of the manuscript:

\begin{manuscript}{Page 5}
Large populations of oscillators are typically described using phase-only models (32), in which the state of each oscillator is represented only by its phase, $\theta$.
\end{manuscript}
\begin{manuscript}{Page 6}
The phase diffusivity parameter $d$ (in units of inverse radians) describes the speed with which the population desynchronizes and can be fit to experimental data (35).
\end{manuscript}
\begin{manuscript}{Page 11}
Previous work has used limit cycle models to estimate population-level parameters, such as desynchronization rate, for phase-only models (35).
\end{manuscript}
\begin{manuscript}{Page 12}
Phase-diffusion models can explain why the gradual damping from experimental population-level data closely resembles an exponentially damped sinusoid, a result that has been shown experimentally (9) and computationally (35).
\end{manuscript}

\begin{reviewer}
Why red and blue are changed to orange in Figure 2D? 
\end{reviewer}

In Fig.~2D, we plot the time-dependent state variables of the two-state model (Model 1).
These variables had not been explicitly plotted in this representation in another figure, so we wished to distinguish them from state-space representations or response curve variables.

\begin{reviewer}
David Rand published a generic theory of response functions which should include ARC?
A comparison is suggested. 
\end{reviewer}

Indeed, David Rand's generic theory of response functions would likely include our amplitude response curve as an example of a differentially changed quantity following a perturbation of finite duration.
Our contribution lies in the definition and derivation of the relevant formulas for amplitude change.
We have included the relevant citation from David Rand, along with a discussion in the manuscript text:

\begin{manuscript}{Page 3}
Mathematical models have long been used to understand the results of circadian experiments (15, 16), aided by definitions and computational techniques designed to match modeling predictions to experimental data.
One such definition is the response function, a general technique that maps a change in an output variable to a temporary change in parameters (17).
For instance, the phase response curve (PRC) has been used to characterize the entrainment behavior of both experimental and mathematical systems (18–20), and in analyzing the synchrony of populations of oscillators (21).
\end{manuscript}

\begin{reviewer}
Joke Meijer (PLoS One) discussed the comparison of single cell amplitudes and ensembles of neuron regarding resetting dynamics as well. 
\end{reviewer}

Thank you for drawing our attention to this work, as it lends some experimental support to the bimodal distributions we observe following a light pulse.
We have incorporated the citation into the relevant section of the manuscript:

\begin{manuscript}{Page 14}
Changes to the synchronization of the population from each perturbation are readily apparent, with both perturbations inducing a bimodal distribution in the phase density.
While experimental data on individual cell phases was unavailable for this data set, bimodal distributions in SCN neuron firing following a phase shift have previously been seen experimentally (49).
As higher-frequency features, these bimodal distributions dissipate as the phases diffuse.
\end{manuscript}

\begin{reviewer}
The critical role of amplitudes for the response to perturbations has been discussed several times in the literature and a discussion of these previous studies can support the relevance of the current manuscript.
Examples: Pittendrigh JBR 1991, RK Barrett J.~Neuroscience 1995, MH Vitaterna PNAS 2006, S Brown PNAS 2008, A Granada PLoS One 2013, A.~Erzberger Proc Royal Soc 2013.

\end{reviewer}

The importance of amplitude in entrainment is an important consideration and has certainly been addressed by many studies in the past.
We have incorporated several of these references into the motivation for the manuscript, as designing therapies to boost circadian amplitudes will surely have an effect on the entrainment of the oscillators as well:

\begin{manuscript}{Page 2}
The amplitude of circadian transcription is a relevant factor, and has been shown to play a critical role in phase resetting and entrainment (3, 4).
Recent studies have further highlighted the importance of high peripheral clock amplitudes in maintaining metabolic health.
\end{manuscript}

\begin{reviewer}
Ute Abraham et al. (Molecular Systems Biology) compare also coupled cells (SCN) with less coupled cells (lung) regarding their entrainment properties and discuss amplitude and relaxation effects.
\end{reviewer}

We have incorporated this reference into two places in the manuscript. First, in the introduction as reference (4) noted above, and also in the conclusion where we discuss the relative entrainment speed of coupled vs. uncoupled cells:

\begin{manuscript}{Page 18}
It is therefore interesting that liver cells have not developed a stronger mechanism of intercellular coupling, such as in the SCN, to maintain robust amplitudes in the absence
of external cues.
Perhaps the ability to quickly re-entrain to a phase shift, which is often slowed by coupling (4), has historically been more advantageous than protection against highly variable food intake.
\end{manuscript}

\end{document}

### 2014-09-02 main_text.tex
fixing error in line wrapping
Adding Latex version of response letter - few changes

@@
 To efficiently capture the population-level effects of bioluminescence experiments, we couple the detailed single-cell ODE model to a phase-density model.
-Previous work has used limit cycle models to estimate population-level parameters, such as desynchronization rate, for phase-only models \cite{Rougemont2007}.
+Previous work has used limit cycle models to estimate population-level parameters, such as desynchronization rate, for phase-only models \cite{Rougemont2008}.
 In this study, we use an ODE model to calculate the response to a perturbation at each phase, and subsequently take the weighted average of these responses according to the phases of the cells in the population.
@@
 To capture this phenomenon {\itshape in silico}, we used a recent model of the core circadian feedback circuit \cite{Hirota2012}.
-In order to find parameters for the phase density function, we assumed an
-initial phase population with $\sigma = 0.1$ and calculated a phase diffusivity
-parameter $d = 0.104$ from the experimental control trajectory.
+In order to find parameters for the phase density function, we assumed an initial phase population with $\sigma = 0.1$ and calculated a phase diffusivity parameter $d = 0.104$ from the experimental control trajectory.
 

### 2014-09-02 response_letter.tex
Updates to response letter, sending to Frank

@@
 The reviewer's point is a good one, however, and we have endeavored to draw as clear a line as possible between new concepts and derivations and those which come from the literature.
-To this end, we have moved as much of the introductory derivations as possible to the ``Materials and Methods'' section, introducing previous work on phase-diffusion models before proceeding with sensitivity-based methods for finding the ARC.
-Additionally, we have included additional references for historical derivations of many of the results.
+{\itshape\bfseries To this end, we have moved as much of the introductory derivations as possible to the ``Materials and Methods'' section}, introducing previous work on phase-diffusion models before proceeding with sensitivity-based methods for finding the ARC.
+Additionally, we have included additional references for historical derivations of many of the results:
+\begin{manuscript}{Page 21}
+\bfseries
+33. Kuramoto, Y., 1984. Chemical Oscillations, Waves, and Turbulence. Springer-Verlag Berlin
+Heidelberg.\\
+34. Teramae, J.-n., and D.~Tanaka, 2004. {Robustness of the noise-induced phase synchronization in a general
+  class of limit cycle oscillators.} \emph{Phys. Rev. Lett.} 93:204103.
+\end{manuscript}
 
@@
 
-While we have left some introductory derivations in the Results section (notably, derivations of the PRC and ARC for phase-only models), these sections have been clearly labeled.
+While we have left some introductory derivations in the Results section (notably, derivations of the PRC and ARC for phase-only models), these sections have been clearly labeled:
+
+\begin{manuscript}{Page 11}
+\bfseries
+We next briefly describe how to calculate the PRC and ARC for a phase-density model.
+These approaches are commonly used in understanding phase models (14, 33), and are presented here to match previous definitions for single cells.
+\end{manuscript}
 
@@
 We have followed the reviewer's suggestion and expanded the section of the manuscript that focuses on the application to experimental data.
-Specifically, we detail our method of selecting the parameter knockdown as well as the initial phase through the use of response curves.
-In doing so, we have expanded Fig.~5 to include response curves for the first parameter pulse, demonstrating the strong dip in the population-level ARC.
+For instance, we detail our method of selecting the parameter knockdown as well as the initial phase through the use of response curves:
+
+\begin{manuscript}{Page 16}
+{\bfseries
+The next step in capturing the experimental data was finding a perturbation capable of desynchronizing the system.
+To find such a perturbation, we calculated population-level ARCs at several different knockdown strengths for each parameter in the model.
+We selected the} degradation rate of {\itshape Per} mRNA {\bfseries from the feasible parameters}, due to PER's known induction by CREB following photo-perturbation (46).
+\end{manuscript}
+
+We have also expanded Fig.~5 to include response curves for the first parameter pulse, demonstrating the strong dip in the population-level ARC:
+
+\begin{manuscript}{Page 17}
+\bfseries
+Figure 5: Method allows direct comparison between model results and experimental bioluminescence profiles.
+({\bfseries A}) Response curves for the parameter perturbation used to model the first light pulse in (B).
+The initial phase of the system is chosen to coincide with the strong dip in the population-level ARC (left). \dots
+\end{manuscript}
 
@@
 We acknowledge that our placement of the introductory material for the phase density methods likely gave the impression that our major results were already well-known.
-We believe our revised manuscript gives a better emphasis to the novel aspects of the study.
+We believe our revised manuscript gives a better emphasis to the novel aspects of the study, which we highlight in the introduction:
+
+\begin{manuscript}{Page 4}
+In this study, {\bfseries we describe approaches} to quantify amplitude change in a population of {\bfseries non-interacting} oscillators.
+{\bfseries By exploiting the independence of each oscillator, we derive computationally efficient methods to approximate the mean dynamics of full stochastic simulations.
+Additionally, our method allows the calculation of ARCs at both the single-cell and population level, allowing the behavior of the system to be quickly profiled.}
+\end{manuscript}
 
@@
 
-We have modified extensive portions of the manuscript in order to make the novel contributions of the manuscript more clear, and added additional explanations to the application of the experimental data.
+We have modified extensive portions of the manuscript in order to make the novel contributions of the manuscript more clear, and added additional explanations to the application of the experimental data:
+
+\begin{manuscript}{Page 16}
+From the experimental system, it appears the light pulse temporarily reduces PER2-luciferase amplitudes immediately following perturbation, separate from the overall change in population synchrony.
+{\bfseries This result is most obvious after the second light pulse, where rhythms require some time before their maximum amplitude is reached.
+Such a delay} is indicative of perturbations to the {\bfseries limit cycle system, suggesting} a contribution from single-cells in determining the population amplitude.
+{\bfseries Light-induced amplitude suppression in fibroblasts is therefore likely mediated at the single-cell level at short time-scales, and at the population level for longer time-scales.
+To improve the fit of the model to the data, the response curves could be fit to match experimentally predicted values.
+Infinitesimal PRCs and ARCs calculated at the single-cell level are numerically efficient to evaluate, and could be more readily incorporated into a parameter estimation algorithm than explicit stochastic simulations of a population of oscillators.}
+\end{manuscript}
+
+Additionally, {\itshape\bfseries many details on the derivations on PRCs have been removed.}
 
@@
 Minor comments: 
-- Equation 11. Could $\Theta(x_0)$ be defined as $\theta$ such that $lim_{t\rightarrow \infty} | x(t)-x^\gamma(\theta+t)|=0$ ?
+- Equation 11.
+Could $\Theta(x_0)$ be defined as $\theta$ such that $lim_{t\rightarrow \infty} | x(t)-x^\gamma(\theta+t)|=0$?
 \end{reviewer}
 
-Yes, although for length constraints (and to remove excess review of results on PRCs) this equation has been removed.
+Yes, although for length constraints (and to remove excess review of results on PRCs) {\itshape\bfseries this equation has been removed.}
 
@@
 
+Yes, and we have changed this definition accordingly:
+
+\begin{manuscript}{Page 7}
+\begin{equation}
+  \mu \coloneqq \int_0^{2\pi} \frac{x^\gamma(\theta)}{2\pi} \; d\theta
+  \tag{21}
+\end{equation}
+\end{manuscript}
+
+Our original definition was intended to demonstrate why $x(t)$ and $y(t)$ would have identical means, but we have simplified the definition to the one suggested.
+
 \begin{reviewer}
@@
 
-We have fixed this regrettable error, and thank the reviewer for their careful reading.
+We have fixed this regrettable error, and thank the reviewer for their careful reading:
+
+\begin{manuscript}{Page 10}
+\begin{equation}
+  \frac{d}{d\tilde{t}}\frac{dA}{dp} \coloneqq \lim_{\tilde{d},\; \Delta p \to 0} \frac{\Delta A}{\tilde{d}\, \Delta p}
+  \tag{31}
+\end{equation}
+\end{manuscript}
 
@@
 \begin{manuscript}{Page 16}
+\bfseries
 The single-cell and population-level response curves for this parameter choice are shown in Fig.~5A, which demonstrates the strong desynchronization that occurs at $\theta\approx\nicefrac{\pi}{4}$.
@@
 This mistake has been corrected.
-Eq.~22 (now 27) is obtained by factoring the difference of squares from the previous equation, but we mistakenly kept a squared term from the previous step.
+Eq.~22 (now 27) is obtained by factoring the difference of squares from the previous equation, but we mistakenly kept a squared term from the previous step:
+
+\begin{manuscript}{Page 10}
+  \begin{equation}
+  = \frac{1}{\delta x(0)} \left[\left(\delta x(\tilde{t}) -
+    \tilde{f}\left(x^\gamma(t_\theta)\right)\delta\theta\right) \left(\delta
+    x(\tilde{t}) + \tilde{f}\left(x^\gamma(t_\theta)\right)\delta\theta +
+    2(x^\gamma(t_\theta) - \mu)\right) \right]
+    \tag{27}
+  \end{equation}
+\end{manuscript}
 
@@
 However, including the delta theta terms explicitly made the equation too long for a single line.
-Since the equation (with the delta theta terms included) was largely a repetition of the previous equation, we have removed it.
+Since the equation (with the delta theta terms included) was largely a repetition of the previous equation, {\bfseries\itshape we have removed it.}
 
@@
 
-This typo was also fixed by removing the squared term in Eq.~24 (now 29), which was unfortunately propagated from the typo in Eq.
-22.
+This typo was also fixed by removing the squared term in Eq.~24 (now 29), which was unfortunately propagated from the typo in Eq.~22.
+
+\begin{manuscript}{page 10}
+  \begin{equation}
+  = 2\left(S(\tilde{t}) - \tilde{f}(x^\gamma(t_\theta))\frac{d\theta}{dx}\right)\left(x^\gamma(t_\theta) - \mu\right)
+    \tag{29}
+  \end{equation}
+\end{manuscript}
 
@@
 
-This error has been fixed.
+{\itshape\bfseries This error has been fixed}, as noted in our response to Reviewer \#1.
 
@@
 \begin{manuscript}{Page 10}
-Here, the first term of the integrand $(S - \tilde{f}\dot{\theta})$ tracks the distance from the perturbed trajectory to the limit cycle, which decays to zero as $t \to \infty$.
-The second term $(x^\gamma - \mu)$ weights this distance by whether or not the deviation occurs above or below the oscillatory mean, yielding negative amplitude changes when the trajectory is perturbed closer to the mean.
-Just as with the parameter-impulse PRC, the infinitesimal parameter-impulse ARC is defined as:
+  {\bfseries Here, the first term of the integrand $(S - \tilde{f}\dot{\theta})$ tracks the distance from the perturbed trajectory to the limit cycle, which decays to zero as $t \to \infty$.
+The second term $(x^\gamma - \mu)$ weights this distance by whether or not the deviation occurs above or below the oscillatory mean, yielding negative amplitude changes when the trajectory is perturbed closer to the mean.}
+Just as with the parameter-impulse PRC, the infinitesimal parameter-impulse ARC {\bfseries is defined as}:
 \begin{equation}
@@
 \end{equation}
-and may be calculated from the state-impulse version with the following relationship:
+{\bfseries and} may be calculated from the state-impulse version with the following relationship:
 \begin{equation}
@@
 \end{equation}
-As with Eq.~11, this equivalency reflects the fact that for a pulse of infinitely short duration, a parameter change is equivalent to changing the state of the system along the direction specified by the Jacobian.
+{\bfseries As with Eq.~11, this equivalency reflects the fact that for a pulse of infinitely short duration, a parameter change is equivalent to changing the state of the system along the direction specified by the Jacobian.}
 \end{manuscript}
@@
 
-The terminology has been changed throughout the manuscript.
+{\bfseries\itshape The terminology has been changed throughout the manuscript.}
 We thank the reviewer for this correction.
@@
 \begin{manuscript}{Page 14}
-In this case, the population-level phase change is a slightly smoothed version of the single-cell PRC, since the population has an averaging effect on incoming perturbations with each cell receiving the input at a slightly different internal phase.
-This smoothing follows from the consistent definition of phase between the single-cell and population level.
-In the ARC plot, however, the shape of the single-cell and population-based ARC are different, since they describe different types of changes (finite vs.\ sustained) in the output trajectories.
+  {\bfseries In this case,} the population-level phase change is a slightly smoothed version of the single-cell {\bfseries PRC}, since the population has an averaging effect on incoming perturbations with each cell receiving the input at a slightly different internal phase.
+{\bfseries This smoothing follows from the consistent definition of phase between the single-cell and population level.
+In the ARC plot, however, the shape of the single-cell and population-based ARC are different, since they describe different types of changes (finite vs.\ sustained) in the output trajectories.}
 \end{manuscript}
@@
 
-This is a general statement regarding the dynamics of populations and limit-cycle systems.
+{\bfseries\itshape This is a general statement} regarding the dynamics of populations and limit-cycle systems.
 Ultimately, non-interacting oscillators have no driving force to re-entrain, so changes to synchrony remain long after limit-cycle oscillations are restored.
@@
 \begin{manuscript}{Page 14}
+  \bfseries
 These examples demonstrate the qualitative differences between amplitude change mediated at the single-cell and population level.
@@
 We therefore do not believe that incorrect noise estimates from the Gillespie algorithm affects our approximations of the stochastic population.
-Regardless, we tested the method with two additional models.
+{\itshape\bfseries Regardless, we tested the method with two additional models.}
 First, the Oregonator, is a simple mass-action oscillatory reaction system considered in Gillespie's 1977 paper.
@@
 
+\begin{manuscript}{Supplemental Info, Page 6}
+\bfseries
+Figure S3: Validation of continuous methods with alternate models.
+(\bfseries A) The Oregonator model (Model 3) was tested as an example of a mass-action limit cycle oscillator.
+A population of 2000 oscillators was perturbed at size different mean phases, $\mu$, by increasing the $\mathit{c2}$ parameter by $50\%$ for $d=\pi$ (highlighted region).
+Only the average $\mathit{Y2}$ variable is plotted.
+({\bfseries B}) The model described in (2) (Model 2) was similarly tested at a variety of mean phases by reducing the $\mathit{vdp}$ parameter by $28.5\%$ and plotting the resulting $\mathit{c2}$ state variable (as in Fig.~5).
+Parameters for the continuous approximations were found by estimating the phase, initial standard deviation, period, and phase diffusivity of the control population.
+Good agreement is seen between the continuous and stochastic simulations, indicating the proposed method is suitable for predicting the population-level responses for a variety of model types.
+\end{manuscript}
+
 \begin{reviewer}
@@
 
-1) In most case, "hat" mark is used to represent various perturbed variables.
-However, "hat" mark is also used to represent the rescaled time variable.
+1) In most case, ``hat'' mark is used to represent various perturbed variables.
+However, ``hat'' mark is also used to represent the rescaled time variable.
 This can lead to confusion in understanding the equations.
@@
 
-This is a good suggestion, and we have changed the notation for scaled time to $\tilde{t}$ throughout the manuscript.
+This is a good suggestion, and {\itshape\bfseries we have changed the notation for scaled time to $\tilde{t}$ throughout the manuscript.}
 Additionally, we noticed that our notation for pulse duration conflicted with our definition of phase diffusivity, so pulse duration has been changed to $\tilde{d}$.
@@
 \begin{manuscript}{Page 5}
-This result follows from the fact that in the limit of an infinitely short and small parameter pulse, $\Delta x(0) \to \nicefrac{d\tilde{f}}{dp}\,\tilde{d}\, \Delta p$.
+  Methods for efficiently calculating these quantities using ODE sensitivity analysis have been developed (24), with the important result that {\bfseries the parameter- and state-impulse PRCs can be related by the Jacobian matrix:}
+\begin{equation}
+  \frac{d}{d\tilde{t}}\frac{d\theta}{dp} = \frac{d\theta}{dx}\frac{d\tilde{f}}{dp} 
+  \tag{11}
+\end{equation}
+{\bfseries This result follows from the fact} that in the limit {\bfseries of an infinitely short and small parameter pulse, $\Delta x(0) \to \nicefrac{d\tilde{f}}{dp}\,\tilde{d}\, \Delta p$.
+This result follows from the fact that in the limit of an infinitely short and small parameter pulse, $\Delta x(0) \to \nicefrac{d\tilde{f}}{dp}\,\tilde{d}\, \Delta p$.}
 \end{manuscript}
@@
 
-This plot has been corrected.
+{\bfseries\itshape This plot has been corrected.}
 
@@
 \begin{manuscript}{Page 14}
-We next demonstrate the usefulness of single-cell and population level ARCs in predicting the mean response of a population.
-Using a detailed single-cell model (Model 2), we calculate the response of the system, specifically the average CRY2 protein expression level, to a temporary increase in {\itshape Per} transcription rate.
-In Fig.~4A, we plot the resulting single-cell and population-level response curves.
+{\bfseries We next demonstrate the usefulness of single-cell and population level ARCs in predicting the mean response of a population.
+  Using a detailed single-cell model (Model 2), we calculate the response of the system, specifically the average CRY2 protein expression level, to a temporary increase in {\itshape Per} transcription rate.}
+In Fig.~4A, we plot {\bfseries the resulting} single-cell and population-level response curves.
 \end{manuscript}
@@
 \begin{manuscript}{Page 15}
-Figure 4: Response curves describe different types of amplitude change.
+  Figure 4: Response curves {\bfseries describe different types of amplitude change
 A population of cells is simulated using a detailed model of circadian rhythms (Model 2).
-Each cell in the population is subjected to a $20\%$ increase in the $\mathit{vtp}$ parameter ({\itshape Per} transcription rate) for $d=\nicefrac{\pi}{2}$, inducing both a phase and amplitude change.
+Each cell in the population is subjected to a $20\%$ increase in the $\mathit{vtp}$ parameter ({\itshape Per} transcription rate) for $d=\nicefrac{\pi}{2}$, inducing both a phase and amplitude change.}
 \end{manuscript}
@@
 The second contribution to the perturbed population trajectory comes from deviations from limit cycle oscillations in each cell.
-We calculate the population-level effect of these deviations by averaging over the deviations that occur at each phase.
+{\bfseries We calculate the population-level effect of these deviations by averaging over the deviations that occur at each phase.}
 We define the deviation trajectory $\delta x(\theta, \tilde{t})$ for each phase as the distance between the perturbed trajectory and the phase-adjusted reference:
@@
 \end{align}
-Since the perturbed trajectory ultimately converges with the phase-adjusted reference, deviations will converge to zero.
+{\bfseries Since the perturbed trajectory ultimately converges with the phase-adjusted reference, deviations will converge to zero.
 Since the phase change, $\Delta\theta$, associated with a perturbation at each phase is likely not known prior to calculating the perturbed trajectory, it is difficult to tabulate deviation trajectories associated with each final phase, $\theta_0 + \Delta\theta$.
-It is therefore more straightforward to find the average effect of single-cell perturbations at the population level by weighting the deviations by the phase density function prior to perturbation.
+It is therefore more straightforward to find the average effect of single-cell perturbations at the population level by weighting the deviations by the phase density function prior to perturbation.}
 \end{manuscript}
@@
 \begin{manuscript}{Page 18}
+\bfseries
 While our examples have focused on light-mediated perturbations, these approaches could also be applied to pharmacological perturbations.
@@
 As medium or temperature changes are often enough to resynchronize cultured cells, a transient application is difficult to achieve experimentally.
-The search for clock-enhancing molecules has therefore tended to focus on constant drug concentrations: for instance, dose-dependent period or amplitude change following inhibition of CK1$\delta$ or similar targets (50).
+The search for clock-enhancing molecules has therefore tended to focus on constant drug concentrations: for instance, dose-dependent period or amplitude change following inhibition of CK1$\delta$ or similar targets (48).
 For {\itshape in vivo} systems, however, pharmacokinetics dictates a finite duration of action for both naturally secreted hormones and pharmacological therapies.
@@
 The decay rate is estimated from stochastic simulations of detailed limit cycle models, and used to capture unperturbed bioluminescence data.
-We have incorporated two relevant references: Rougement et al, 2006 and Rougement et al, 2007, and added the references to the following sections of the manuscript:
+We have incorporated two relevant references:
+
+\begin{manuscript}{Page 21}
+{\bfseries
+32. Rougemont, J., and F. Naef, 2006. Collective synchronization in populations
+of globally coupled phase oscillators with drifting frequencies. {\itshape Phys.\ Rev.\ E} 73:011104.\\}
+\dots\\
+{\bfseries
+  36. Rougemont, J., and F. Naef, 2007. Dynamical signatures of cellular fluctuations and oscillator stability in peripheral circadian clocks. {\itshape Mol.\ Syst.\ Biol.\ } 3:93. }
+\end{manuscript}
+
+and added the references to the following sections of the manuscript:
 
 \begin{manuscript}{Page 5}
+\bfseries
 Large populations of oscillators are typically described using phase-only models (32), in which the state of each oscillator is represented only by its phase, $\theta$.
@@
 \begin{manuscript}{Page 6}
-The phase diffusivity parameter $d$ (in units of inverse radians) describes the speed with which the population desynchronizes and can be fit to experimental data (35).
+The phase diffusivity parameter $d$ (in units of inverse radians) describes the speed with which the population desynchronizes and can be fit to experimental data {\bfseries (36)}.
 \end{manuscript}
 \begin{manuscript}{Page 11}
-Previous work has used limit cycle models to estimate population-level parameters, such as desynchronization rate, for phase-only models (35).
+\bfseries
+Previous work has used limit cycle models to estimate population-level parameters, such as desynchronization rate, for phase-only models (36).
 \end{manuscript}
 \begin{manuscript}{Page 12}
-Phase-diffusion models can explain why the gradual damping from experimental population-level data closely resembles an exponentially damped sinusoid, a result that has been shown experimentally (9) and computationally (35).
+  Phase-diffusion {\bfseries models can explain} why the gradual damping from experimental population-level data closely resembles an exponentially damped sinusoid, {\bfseries a result that has been shown experimentally (9) and computationally (36).}
 \end{manuscript}
@@
 \begin{manuscript}{Page 3}
-Mathematical models have long been used to understand the results of circadian experiments (15, 16), aided by definitions and computational techniques designed to match modeling predictions to experimental data.
-One such definition is the response function, a general technique that maps a change in an output variable to a temporary change in parameters (17).
-For instance, the phase response curve (PRC) has been used to characterize the entrainment behavior of both experimental and mathematical systems (18–20), and in analyzing the synchrony of populations of oscillators (21).
+Mathematical models have long been used to understand the results of circadian experiments (15, 16), {\bfseries aided by} definitions and computational techniques {\bfseries designed to match modeling predictions to experimental data.
+One such definition is the response function, a general technique that maps a change in an output variable to a temporary change in parameters (17).}
+For instance, the phase response curve (PRC) has been used to characterize the entrainment behavior of both experimental and mathematical systems (18-20), and in analyzing the synchrony of populations of oscillators (21).
 \end{manuscript}
@@
 \begin{manuscript}{Page 14}
-Changes to the synchronization of the population from each perturbation are readily apparent, with both perturbations inducing a bimodal distribution in the phase density.
-While experimental data on individual cell phases was unavailable for this data set, bimodal distributions in SCN neuron firing following a phase shift have previously been seen experimentally (49).
-As higher-frequency features, these bimodal distributions dissipate as the phases diffuse.
+Changes to the synchronization of the population from each perturbation are readily apparent, {\bfseries with both perturbations inducing a bimodal distribution in the phase density.
+While experimental data on individual cell phases was unavailable for this data set, bimodal distributions in SCN neuron firing following a phase shift have previously been seen experimentally (47).
+As higher-frequency features, these bimodal distributions} dissipate as the phases diffuse.
 \end{manuscript}
@@
 \begin{manuscript}{Page 2}
-The amplitude of circadian transcription is a relevant factor, and has been shown to play a critical role in phase resetting and entrainment (3, 4).
+  {\bfseries The amplitude of circadian transcription is a relevant factor, and has been shown to play a critical role in phase resetting and entrainment (3, 4).}
 Recent studies have further highlighted the importance of high peripheral clock amplitudes in maintaining metabolic health.
@@
 of external cues.
-Perhaps the ability to quickly re-entrain to a phase shift, which is often slowed by coupling (4), has historically been more advantageous than protection against highly variable food intake.
+Perhaps the ability to quickly re-entrain to a phase shift, which is often slowed by coupling {\bfseries (4)}, has historically been more advantageous than protection against highly variable food intake.
 \end{manuscript}

### 2014-09-02 main_text.tex
Updates to response letter, sending to Frank

--- 2014-08-29 main_text.tex
@@
 The shape of $p(\theta, \tilde{t})$ changes as the cells advance in time.
-Stochastic effects cause the population to gradually desynchronize as slight cycle-to-cycle deviations are propagated throughout the population.
+Stochastic effects cause the population to gradually desynchronize as slight cycle-to-cycle deviations are propagated throughout the population \cite{Teramae2004}.
 For a infinite population of oscillators, these effects are well-described by a Fokker-Plank equation \cite{Stein1965}:
@@
 \begin{equation}
-  \frac{d}{d\tilde{t}}\frac{dA}{dp} \coloneqq \lim_{\tilde{d},\; \Delta p \to 0}
-  \frac{\Delta A}{\tilde{d}\, \Delta p}
+  \frac{d}{d\tilde{t}}\frac{dA}{dp} \coloneqq \lim_{\tilde{d},\; \Delta p \to 0} \frac{\Delta A}{\tilde{d}\, \Delta p}
 \end{equation}
@@
 \bibliographystyle{biophysj.bst}
+% \bibliography{library.bib}
 \bibliography{condensed_library.bib}

### 2014-09-02 response_letter.tex
Updating the diff and response letter, (actually) sent to Frank

@@
 {\itshape\bfseries To this end, we have moved as much of the introductory derivations as possible to the ``Materials and Methods'' section}, introducing previous work on phase-diffusion models before proceeding with sensitivity-based methods for finding the ARC.
-Additionally, we have included additional references for historical derivations of many of the results:
+Additionally, we have included additional references for historical derivations of the results:
+\begin{manuscript}{Pages 5-6}
+  The synchrony of the population can be modeled using a probability density function $p(\theta, \tilde{t})$ that describes the probability of finding an oscillator at each phase {\bfseries (33)}.
+The usefulness of probability density functions in describing the phase and amplitude responses of populations of circadian cells has been {\bfseries previously shown} (14).
+As with all probability density functions:
+\begin{equation}
+  \int_0^{2\pi} p(\theta, \tilde{t}) \; d\theta = 1
+  \tag{12}
+\end{equation}
+The shape of $p(\theta, \tilde{t})$ changes as the cells advance in time.
+Stochastic effects cause the population to gradually desynchronize as slight cycle-to-cycle deviations are propagated throughout the population {\bfseries (34).
+For a infinite population of oscillators,} these effects are well-described by a {\bfseries Fokker-Plank} equation (35):
+\end{manuscript}
+
 \begin{manuscript}{Page 21}

### 2014-09-04 main_text.tex
Adding ackowledgement to anonymous reviewers

@@
 \section*{Acknowledgments}
-We thank Yongqiang Wang for the helpful discussions.
-This work was supported by the National Institutes of Health/National Institute of General Medical Sciences under award number 1R01GM096873-01 and by the Institute for Collaborative Biotechnologies through grant W911NF-09-0001 from the U.S.
-Army Research Office.
+We thank Yongqiang Wang for the helpful discussions, and gratefully acknowledge the three anonymous reviewers whose comments/suggestions helped improve this work. 
+This work was supported by the National Institutes of Health/National Institute of General Medical Sciences under award number 1R01GM096873-01 and by the Institute for Collaborative Biotechnologies through grant W911NF-09-0001 from the U.S.\ Army Research Office.
 

### 2014-09-25 response_letter.tex
starting second revision, adding changes requested by first reviewer

--- 2014-09-02 response_letter.tex
@@
 
-Thank you for your email on August 1st, 2014, inviting us to submit a revised version of our manuscript, ``Amplitude metrics for cellular circadian bioluminescence reporters'' (MS: 2014BIOPHYSJ304278) by Peter C.
+Thank you for your email on September 24th, 2014, inviting us to submit a revised version of our manuscript, ``Amplitude metrics for cellular circadian bioluminescence reporters'' (MS: 2014BIOPHYSJ304278R) by Peter C.
 St.
@@
 
-We thank the reviewers for their detailed comments of the manuscript and suggestions for how to improve the work.
-We did our best to balance the suggestions of Reviewer \#1, who suggested trimming much of the theoretical foundations of the method, with the suggestions of Reviewer \#2, who recommended additional explanations and validations.
-Ultimately we believe that the changes have greatly strengthened our manuscript by demonstrating our method is generalizable to alternative models and by focusing the work on our novel contributions.
+We thank the reviewers for their second round of comments of the manuscript, in particular Reviewer \#2 for their continued attention to detail.
+We believe addressing these remaining concerns and suggestions have improved the polish of the manuscript.
 
@@
 \begin{reviewer}
-In this manuscript the authors study the amplitude of oscillators with a focus on circadian rhythms.
-Understanding the effect of perturbations of populations of oscillators is important to understand the mechanisms of circadian rhythms.
+The authors have largely responded you my concerns.
 
-The authors start by describing phase response curves PRCs, ARCs and phase density equations.
-They also describe practical approaches to computing these.
-Using such a computed perturbed distribution, the authors show how to calculate an ``average oscillation'' of the population.
+Typos and suggestions
 
-My main concern is that I cannot tell what is novel in the manuscript.
-Since references are scant in the main body of the text, this is somewhat obscure.
-For example, phase response curves have been discussed in numerous places, as have amplitude response curves.
-The phase density methods and an associated Fokker-Planck equation (Eq.~31) also date back at least to Kuramoto and Stratonovich, with a lot of work following that.
-For instance, the ``synchronization index'', $\rho$, is the order parameter in Kuramoto's treatment, and has been studied in hundreds of papers.
-I guess this is well known, but some references should be given.
-Stochastic oscillators at the population level, with a focus on PRCs, have been studied in a series of papers by Tanaka, Goldobin, Ermentrout and others - see also the standard, but somewhat dated book by Pikovsky, Rosenblum and Kurths.
-
-Pages 4-14 of the manuscript therefore contain a lot of known results.
-Some of the results may be new - for instance, I have not seen Eq.~(26) before - but the majority of the results presented here have been standard fare in the voluminous literature on stochastic oscillators.
-Is the novelty here that these can be used to analyze data? If so, this is not well explained.
+- p. 2: "fail to capture the collective dynamics of a population oscillators" 
 \end{reviewer}
 
-We acknowledge that many of the mathematical formulas presented in the ``Results'' section have been used by many other studies: our intention was to introduce concepts and notations to readers who might be less familiar with the large body of previous literature.
-The reviewer's point is a good one, however, and we have endeavored to draw as clear a line as possible between new concepts and derivations and those which come from the literature.
-To this end, we have moved as much of the introductory derivations as possible to the ``Materials and Methods'' section, introducing previous work on phase-diffusion models before proceeding with sensitivity-based methods for finding the ARC.
-Additionally, we have included additional references for historical derivations of many of the results.
+This typo has been addressed, the sentence now reads:
 
-To address the point directly, we consider the following contributions of the manuscript as novel, and highlight these in the results section:
-
-\begin{itemize}
-  \item Our definition of transient amplitude change (at the single-cell level).
-This definition allows us to calculate and directly compare amplitude change using both finite-difference and ODE sensitivity methods.
-  \item The subsequent derivation of the differential ARC, similar to derivations of the differential phase response curve.
-  \item We derive a method to approximate a stochastic population of noninteracting oscillators using continuous methods, including accounting for deviations from limit cycle trajectories.
-\end{itemize}
-
-While we have left some introductory derivations in the Results section (notably, derivations of the PRC and ARC for phase-only models), these sections have been clearly labeled.
-
-\begin{reviewer}
-The interesting part of the manuscript are the application to experimental data.
-However, this part of the manuscript is very brief.
-It is also not clear to me what methods discussed in the first part of the manuscript were used here.
-An ARC was computed, and the phase density is also found for the model.
-However, details on the connection to the first part of the manuscript are sparse.
-\end{reviewer}
-
-We have followed the reviewer's suggestion and expanded the section of the manuscript that focuses on the application to experimental data.
-Specifically, we detail our method of selecting the parameter knockdown as well as the initial phase through the use of response curves.
-In doing so, we have expanded Fig.~5 to include response curves for the first parameter pulse, demonstrating the strong dip in the population-level ARC.
-
-\begin{reviewer}
-Ultimately, I am not sure what the point of the manuscript is. The authors state that
-
-``\dots we provide new tools and definitions to quantify amplitude change in a population of oscillators, and determine what factors dominate amplitude change under certain conditions.''
-
-Then I disagree - most of the tools here are not new.
-Density methods for uncoupled populations of oscillators have been around for many years - at least since Stratonovich.
-What is novel is potentially the treatment of the amplitude response curve.
-If this is indeed the novel contribution, then this needs to be better explained.
-\end{reviewer}
-
-We acknowledge that our placement of the introductory material for the phase density methods likely gave the impression that our major results were already well-known.
-We believe our revised manuscript gives a better emphasis to the novel aspects of the study.
-
-\begin{reviewer}
-I suggest rewriting the manuscript with an emphasis on the application.
-It is still a bit unclear how the proposed method can be applied to real data, and what one would get out of it.
-Therefore a focus on these ideas, rather than the review of results about PRCs, ARCs and phase densities, would significantly improve the manuscript.
-\end{reviewer}
-
-We have modified extensive portions of the manuscript in order to make the novel contributions of the manuscript more clear, and added additional explanations to the application of the experimental data.
-
-\begin{reviewer}
-Minor comments: 
-- Equation 11. Could $\Theta(x_0)$ be defined as $\theta$ such that $lim_{t\rightarrow \infty} | x(t)-x^\gamma(\theta+t)|=0$ ?
-\end{reviewer}
-
-Yes, although for length constraints (and to remove excess review of results on PRCs) this equation has been removed.
-
-\begin{reviewer}
-- Equation 16: Could $\mu$ just be defined as $\int_0^{2\pi} y(t)dt/(2\pi)$ ?
-\end{reviewer}
-
-\begin{reviewer}
-- Equation 27 is incomplete. 
-\end{reviewer}
-
-We have fixed this regrettable error, and thank the reviewer for their careful reading.
-
-\begin{reviewer}
-- In Figure 5A, the initial condition for the experiments is 1. Why did the authors use the initial condition $0.5$ in the simulations?
-\end{reviewer}
-
-The initial conditions of the experiments likely arises from a normalization of the bioluminescence data, and absolute values were not available to directly compare amplitudes.
-Therefore the relative amplitudes of both systems had to be compared, with the amplitude estimated from the control trajectory.
-The phase offset of the experimental and model systems was also not fit exactly - we added an explanation of the rationale behind this decision to the manuscript:
-
-\begin{manuscript}{Page 16}
-The single-cell and population-level response curves for this parameter choice are shown in Fig.~5A, which demonstrates the strong desynchronization that occurs at $\theta\approx\nicefrac{\pi}{4}$.
-Since the reporter used in the experimental system likely has a phase lag from the corresponding mRNA, we did not try to match the initial phase of the simulation to experiment.
-Instead, the initial phase of simulation was chosen such that the first pulse occurred when the system was at the phase that corresponded with the minimum of the population-level ARC.
-\end{manuscript}
-
-\section*{Reviewer \#2}
-
-\begin{reviewer}
-The manuscript titled ``Amplitude metrics for cellular circadian bioluminescence reporters'' investigates a very interesting problem: amplitude perturbations via changes in single-cell levels and/or population synchrony.
-Furthermore, the manuscript suggests that amplitude changes mediated by single-cell level or population-level changes can be distinguished based on the analysis in the manuscript.
-I totally agree that this is a very important problem in circadian clocks as well as many other biological oscillators.
-However, unlike the statement in the abstract of the manuscript ``In particular, our analysis reveals that amplitude changes mediated by individual cells or population synchrony can be distinguished in tissue-level bioluminescence data without the need for single-cell measurements'', I am not convinced how their analysis provides such a tool.
-Furthermore, I found some mistakes in their derivation of a major formula, which reduces the reliability of the manuscript.
-Finally, I am also concerned about major simulation results because the Gillespie algorithm with non-elementary propensity functions is used, which can cause a large error.
-In summary, I strongly agree that this manuscript considers an interesting and important problem, but due to the above reasons, I am not convinced with their results.
-To resolve these issues I have the following suggestions:
-
-Major concerns
-
-1) During the derivation of the major formula (Eq 18- 28), I found some errors. Please, fix them.
-\end{reviewer}
-
-We thank the reviewer for their careful review of the mathematical derivations. The errors are largely typographical:
-
-\begin{reviewer}
-a. (21) and (22) are not equal. 
-\end{reviewer}
-
-This mistake has been corrected.
-Eq.~22 (now 27) is obtained by factoring the difference of squares from the previous equation, but we mistakenly kept a squared term from the previous step.
-
-\begin{reviewer}
-b. (22) and (23) are not equal. (23) is an approximation of (22) when delta theta approaches to zero. 
-\end{reviewer}
-
-This is a fair point, as we dropped the infinitesimal elements before taking the limit.
-The equality is valid in the differential limit as the phase shift for an infinitesimal perturbation will be identically 0, and therefore this error in notation does not affect the validity of the results.
-However, including the delta theta terms explicitly made the equation too long for a single line.
-Since the equation (with the delta theta terms included) was largely a repetition of the previous equation, we have removed it.
-
-\begin{reviewer}
-c. (23) and (24) are not equal.
-\end{reviewer}
-
-This typo was also fixed by removing the squared term in Eq.~24 (now 29), which was unfortunately propagated from the typo in Eq.
-22.
-
-\begin{reviewer}
-d. (27) is not complete.
-\end{reviewer}
-
-This error has been fixed.
-
-\begin{reviewer}
-e. Please, provide an intuitive explanation for formulas (26) and (28) for the reader. 
-\end{reviewer}
-
-The following two explanatory remarks have been added following Eq.~26 and 28 (now 30 and 32):
-
-\begin{manuscript}{Page 10}
-Here, the first term of the integrand $(S - \tilde{f}\dot{\theta})$ tracks the distance from the perturbed trajectory to the limit cycle, which decays to zero as $t \to \infty$.
-The second term $(x^\gamma - \mu)$ weights this distance by whether or not the deviation occurs above or below the oscillatory mean, yielding negative amplitude changes when the trajectory is perturbed closer to the mean.
-Just as with the parameter-impulse PRC, the infinitesimal parameter-impulse ARC is defined as:
-\begin{equation}
-  \frac{d}{d\tilde{t}}\frac{dA}{dp} \coloneqq \lim_{\tilde{d},\; \Delta p \to 0}
-  \frac{\Delta A}{\tilde{d}\, \Delta p}
-  \tag{31}
-\end{equation}
-and may be calculated from the state-impulse version with the following relationship:
-\begin{equation}
-  \frac{d}{d\tilde{t}}\frac{dA}{dp} = \frac{dA}{dx}\frac{d\tilde{f}}{dp}
-  \tag{32}
-\end{equation}
-As with Eq.~11, this equivalency reflects the fact that for a pulse of infinitely short duration, a parameter change is equivalent to changing the state of the system along the direction specified by the Jacobian.
+\begin{manuscript}{Page 2}
+  Ordinary differential equation (ODE) models of gene regulation are capable of describing the amplitude and phase-resetting behavior of single cells, but fail to capture the collective dynamics of a population {\bfseries of} oscillators.
 \end{manuscript}
@@
 \begin{reviewer}
-2) Whereas the manuscript describes that they explore ``weakly coupled oscillators'', I found that what the manuscript is really considering are ``independent or non-interacting oscillators''.
-Weakly coupled oscillators have been used in a different context (e.g.\ Kuramoto oscillators).
-Please, correct this.
+- mu is defined a few lines after it is first used in Eq. (20). Perhaps define it first? 
 \end{reviewer}
 
-The terminology has been changed throughout the manuscript.
-We thank the reviewer for this correction.
+This change has also been made.
 
-\begin{reviewer}
-3) One of the most interesting results of this manuscript (Fig4) is showing that ARCs, but not PRCs are very different at the single-cell level as compared to the population level.
-I can see that a single simulation (Fig.~4A) supports this, however, I cannot see that the analysis in the manuscript is sufficient to support this statement.
-That is, please explain how the analysis of this manuscript (perhaps the formula for the ARC) leads to the conclusion that appears in Fig.~4A.
-Furthermore, I wonder if the result in Fig.~4A is a specific example of a special case or a general property.
-In particular, a recent study (DeWoskin et al (2014)) shows that PRCs can also be very different in single-cell and population levels.
-To test the generality of the results (Fig.~4A), please test this argument, at least with more simple models such as Model 1 in the manuscript, a Goodwin oscillator based on a Hill function, or the Kim-Forger model based on protein sequestration.
-\end{reviewer}
-
-Our point in discussing the phase response curves of Fig.~4A is only that phase maintains a consistent definition between the population and individual cell level, while amplitude must be defined differently in each case.
-We have added the following text to the manuscript to make this point more explicit:
-
-\begin{manuscript}{Page 14}
-In this case, the population-level phase change is a slightly smoothed version of the single-cell PRC, since the population has an averaging effect on incoming perturbations with each cell receiving the input at a slightly different internal phase.
-This smoothing follows from the consistent definition of phase between the single-cell and population level.
-In the ARC plot, however, the shape of the single-cell and population-based ARC are different, since they describe different types of changes (finite vs.\ sustained) in the output trajectories.
-\end{manuscript}
-
-The PRC result in Fig.~4A is not a special case, but other types of phase-resetting behavior in a population can occur.
-For instance, an almost desynchronized population would undergo type 0 (strong) phase resetting, even if the PRC of each cell was type 1 (weak), since any change to the phase-density profile would create a new mean phase.
-Ultimately, we feel that the literature on the collective phase-response of a population of oscillators (coupled or non-coupled, simulated either explicitly or with phase-density methods) is relatively mature, and that a providing a detailed analysis of the phase responses we obtain is beyond the scope of our major contribution (outlined in response to Reviewer 1).
-
-\begin{reviewer}
-4) The author suggests that ``amplitude change induced at the population level will follow a consistent decaying sinusoid, while a change induced at the single-cell level will deviate from this exponential decay''.
-Is this a general statement or the conclusion of a specific simulation result (Fig.~4B and C)?
-If this is not a general statement, then I doubt whether this result can be used to distinguish two sources of amplitude change without explicit recording of single cell amplitudes.
-If this is a general statement, please provide more details and intuitive explanations, as well as further simulation results with more types of models, or analysis.
-\end{reviewer}
-
-This is a general statement regarding the dynamics of populations and limit-cycle systems.
-Ultimately, non-interacting oscillators have no driving force to re-entrain, so changes to synchrony remain long after limit-cycle oscillations are restored.
-
-In addition to adding simulation results with additional models (see below), we have changed the quoted text to be more explicit in this explanation:
-
-\begin{manuscript}{Page 14}
-These examples demonstrate the qualitative differences between amplitude change mediated at the single-cell and population level.
-Their characteristic features allow us to distinguish between these two sources of amplitude change without explicitly recording single-cell amplitudes: amplitude change induced at the population-level will be sustained, but may be masked in the short term by changes to single-cell level amplitudes.
-Similarly, a change induced at the single-cell level will be evident only at short time-scales.
-These results underscore the importance in considering both single-cell and population-mediated amplitude change in predicting the effects of daily stimuli on clock amplitudes.
-\end{manuscript}
-
-\begin{reviewer}
-5) To derive the major formula in Eq.~(50), various approximations and assumptions were made.
-This formula is tested with a single simulation.
-To confirm the accuracy of Eq (50), please, test this in other models as suggested above.
-\end{reviewer}
-
-(Answered below)
-
-\begin{reviewer}
-6) Stochastic simulations of the model with non-elementary reactions (e.g.
-Hill-functions or Michaelis-Menten terms) are done with the Gillespie algorithm.
-However, recent studies have shown the inaccuracy of this approach (Thomas et al.
-BMC Syst Biol (2012) and Agarwal et al.\ J Chem Phy (2012)).
-Please, discuss whether this inaccuracy in the stochastic simulations affects the simulation results in Fig.~5.
-If possible, please try a model based on mass-action kinetics or a model based on total QSSA since Yao et al.\ Biophy J (2008) shows that the model based on total QSSA can provide accurate simulations with the Gillespie algorithm.
-\end{reviewer}
-
-Points 5 and 6 are linked concerns regarding the simulation and approximation of stochastic trajectories, so we have addressed them together.
-The noise generated using the Gillespie algorithm for non-elementary propensity functions can indeed be different than the noise that would occur for the full system when expanded into elementary reactions.
-That being said, the level of noise present in Fig.~5 was chosen by varying the volume parameter of the stochastic simulation so that desynchronization rates matched experimentally observed data.
-We therefore do not believe that incorrect noise estimates from the Gillespie algorithm affects our approximations of the stochastic population.
-Regardless, we tested the method with two additional models.
-First, the Oregonator, is a simple mass-action oscillatory reaction system considered in Gillespie's 1977 paper.
-We used a perturbation to the conversion rate of two of the intermediate species at 6 different phases to test the accuracy of the perturbed approximation, $x_hat$.
-Second, we tested the method with a stochastic simulation of a more detailed circadian model.
-For consistency, we used the same model and perturbation as was used in Fig.~5 (for the first light pulse).
-The results of these simulations are included as Fig.~S3, which (similar to Fig.~S2) demonstrates good agreement between the stochastic models and continuous approximations.
-
-\begin{reviewer}
-Minor concerns:
-
-1) In most case, "hat" mark is used to represent various perturbed variables.
-However, "hat" mark is also used to represent the rescaled time variable.
-This can lead to confusion in understanding the equations.
-Please, use different notation for scaled time.
-\end{reviewer}
-
-This is a good suggestion, and we have changed the notation for scaled time to $\tilde{t}$ throughout the manuscript.
-Additionally, we noticed that our notation for pulse duration conflicted with our definition of phase diffusivity, so pulse duration has been changed to $\tilde{d}$.
-
-\begin{reviewer}
-2) Please provide a little more detail for the derivation of (14).
-\end{reviewer}
-
-The derivation for Eq.~14 (now 11) is analogous to Eq.~28 (now 32), a lengthy process that was detailed in the appendix of Taylor et al, 2008 (PMID 19593456).
-For space constraints, we have added only brief remarks on their derivation:
-
-\begin{manuscript}{Page 5}
-This result follows from the fact that in the limit of an infinitely short and small parameter pulse, $\Delta x(0) \to \nicefrac{d\tilde{f}}{dp}\,\tilde{d}\, \Delta p$.
-\end{manuscript}
-
-\begin{reviewer}
-3) In Fig.~2A, a discontinuous PRC appears as continuous curve.
-Please, do not link the discontinuous points, and redraw the PRCs.
-\end{reviewer}
-
-This plot has been corrected.
-
-\begin{reviewer}
-4) The PRC and ARC in Fig 4A are in response to which perturbation?
-Please provide details on how Fig 4A is simulated.
-\end{reviewer}
-
-Additional details on the simulation of Fig. 4 have been added to both the main text and the figure legend:
-
-\begin{manuscript}{Page 14}
-We next demonstrate the usefulness of single-cell and population level ARCs in predicting the mean response of a population.
-Using a detailed single-cell model (Model 2), we calculate the response of the system, specifically the average CRY2 protein expression level, to a temporary increase in {\itshape Per} transcription rate.
-In Fig.~4A, we plot the resulting single-cell and population-level response curves.
-\end{manuscript}
-
-\begin{manuscript}{Page 15}
-Figure 4: Response curves describe different types of amplitude change.
-A population of cells is simulated using a detailed model of circadian rhythms (Model 2).
-Each cell in the population is subjected to a $20\%$ increase in the $\mathit{vtp}$ parameter ({\itshape Per} transcription rate) for $d=\nicefrac{\pi}{2}$, inducing both a phase and amplitude change.
-\end{manuscript}
-
-\begin{reviewer}
-5) The definition of Eq (48) is not clear.
-Please provide a more detailed explanation.
-\end{reviewer}
-
-Eq~ 48 (now 44) tracks the distance between the perturbed trajectory and the limit cycle for each phase.
-The weighted-average of this distance is then used to find the population-level expression following perturbation.
-To make this point more clear, we have added additional explanatory text to the manuscript:
-
-\begin{manuscript}{Pages 12-13}
-The second contribution to the perturbed population trajectory comes from deviations from limit cycle oscillations in each cell.
-We calculate the population-level effect of these deviations by averaging over the deviations that occur at each phase.
-We define the deviation trajectory $\delta x(\theta, \tilde{t})$ for each phase as the distance between the perturbed trajectory and the phase-adjusted reference:
-\begin{align}
-  \delta x(\theta_0, \tilde{t}) &\coloneqq x(\tilde{t}) - x^\gamma(\tilde{t} + \theta_0 + \Delta \theta)\tag{48} \\
-  \therefore\; \lim_{\tilde{t} \to \infty} \delta x(\theta, \tilde{t}) &= 0 \tag{49}
-\end{align}
-Since the perturbed trajectory ultimately converges with the phase-adjusted reference, deviations will converge to zero.
-Since the phase change, $\Delta\theta$, associated with a perturbation at each phase is likely not known prior to calculating the perturbed trajectory, it is difficult to tabulate deviation trajectories associated with each final phase, $\theta_0 + \Delta\theta$.
-It is therefore more straightforward to find the average effect of single-cell perturbations at the population level by weighting the deviations by the phase density function prior to perturbation.
-\end{manuscript}
-
-\begin{reviewer}
-6) In a couple of places, the manuscript discusses pharmacological manipulation of circadian clocks.
-Could you discuss this in more detail since the manuscript analyzes data coming from only light-induced perturbations (Fig.~5)? 
-For instance, the manipulation of circadian rhythms with the CK1d/e inhibitor (PF-670462) has been studied widely.
-Can you discuss how your method can be applied to the data in these studies?
-\end{reviewer}
-
-Indeed, while we would have preferred to analyze data coming from pharmacological perturbations, this type of data is difficult to obtain.
-We have added an additional paragraph to the conclusion to address this limitation, and point out potential future applications of the method:
-
-\begin{manuscript}{Page 18}
-While our examples have focused on light-mediated perturbations, these approaches could also be applied to pharmacological perturbations.
-Some difficulty arises in obtaining such data in cultured cells, however, as entraining perturbations would require that pharmacological agents be introduced for only a finite duration.
-As medium or temperature changes are often enough to resynchronize cultured cells, a transient application is difficult to achieve experimentally.
-The search for clock-enhancing molecules has therefore tended to focus on constant drug concentrations: for instance, dose-dependent period or amplitude change following inhibition of CK1$\delta$ or similar targets (50).
-For {\itshape in vivo} systems, however, pharmacokinetics dictates a finite duration of action for both naturally secreted hormones and pharmacological therapies.
-More effective treatments might therefore be designed by explicitly accounting for such a transient response, perturbing peripheral clocks at the right phase to induce resynchronization and an increase in single-cell level amplitudes.
-\end{manuscript}
-
-\section*{Reviewer \#3}
-
-\begin{reviewer}
-Many circadian clock studies use reporters and analyze ensembles of cells.
-Frequently phase shifts are analyzed using Phase Response Curves.
-The authors complement this approach by introducing Amplitude Response Curves.
-Moreover, they discuss effects of ensemble averaging.
-A clear mathematical description is provided which is useful to interpret experimental data.
-Some related studied should discussed to embed the manuscript and to highlight the novelty.
-
-\end{reviewer}
-
-We thank the reviewer for their positive impression of the manuscript, as well as for their helpful suggestions of related studies to discuss.
-We have incorporated many of the suggested works, which we feel helps to place the manuscript within the larger field of circadian modeling.
-
-\begin{reviewer}
-Detailed comments:
-
-Felix Naef studied stochastic models to discuss fibroblast rhythms.
-He found as well purely exponential decay in case of damped oscillators and deviations for limit cycles.
-The current study should be compared to these models.
-\end{reviewer}
-
-In the relevant studies from Felix Naef, the authors use a phase model in which Gaussian white noise causes the frequencies to drift in time.
-The decay rate is estimated from stochastic simulations of detailed limit cycle models, and used to capture unperturbed bioluminescence data.
-We have incorporated two relevant references: Rougement et al, 2006 and Rougement et al, 2007, and added the references to the following sections of the manuscript:
-
-\begin{manuscript}{Page 5}
-Large populations of oscillators are typically described using phase-only models (32), in which the state of each oscillator is represented only by its phase, $\theta$.
-\end{manuscript}
-\begin{manuscript}{Page 6}
-The phase diffusivity parameter $d$ (in units of inverse radians) describes the speed with which the population desynchronizes and can be fit to experimental data (35).
-\end{manuscript}
-\begin{manuscript}{Page 11}
-Previous work has used limit cycle models to estimate population-level parameters, such as desynchronization rate, for phase-only models (35).
-\end{manuscript}
-\begin{manuscript}{Page 12}
-Phase-diffusion models can explain why the gradual damping from experimental population-level data closely resembles an exponentially damped sinusoid, a result that has been shown experimentally (9) and computationally (35).
-\end{manuscript}
-
-\begin{reviewer}
-Why red and blue are changed to orange in Figure 2D? 
-\end{reviewer}
-
-In Fig.~2D, we plot the time-dependent state variables of the two-state model (Model 1).
-These variables had not been explicitly plotted in this representation in another figure, so we wished to distinguish them from state-space representations or response curve variables.
-
-\begin{reviewer}
-David Rand published a generic theory of response functions which should include ARC?
-A comparison is suggested. 
-\end{reviewer}
-
-Indeed, David Rand's generic theory of response functions would likely include our amplitude response curve as an example of a differentially changed quantity following a perturbation of finite duration.
-Our contribution lies in the definition and derivation of the relevant formulas for amplitude change.
-We have included the relevant citation from David Rand, along with a discussion in the manuscript text:
-
-\begin{manuscript}{Page 3}
-Mathematical models have long been used to understand the results of circadian experiments (15, 16), aided by definitions and computational techniques designed to match modeling predictions to experimental data.
-One such definition is the response function, a general technique that maps a change in an output variable to a temporary change in parameters (17).
-For instance, the phase response curve (PRC) has been used to characterize the entrainment behavior of both experimental and mathematical systems (18–20), and in analyzing the synchrony of populations of oscillators (21).
-\end{manuscript}
-
-\begin{reviewer}
-Joke Meijer (PLoS One) discussed the comparison of single cell amplitudes and ensembles of neuron regarding resetting dynamics as well. 
-\end{reviewer}
-
-Thank you for drawing our attention to this work, as it lends some experimental support to the bimodal distributions we observe following a light pulse.
-We have incorporated the citation into the relevant section of the manuscript:
-
-\begin{manuscript}{Page 14}
-Changes to the synchronization of the population from each perturbation are readily apparent, with both perturbations inducing a bimodal distribution in the phase density.
-While experimental data on individual cell phases was unavailable for this data set, bimodal distributions in SCN neuron firing following a phase shift have previously been seen experimentally (49).
-As higher-frequency features, these bimodal distributions dissipate as the phases diffuse.
-\end{manuscript}
-
-\begin{reviewer}
-The critical role of amplitudes for the response to perturbations has been discussed several times in the literature and a discussion of these previous studies can support the relevance of the current manuscript.
-Examples: Pittendrigh JBR 1991, RK Barrett J.~Neuroscience 1995, MH Vitaterna PNAS 2006, S Brown PNAS 2008, A Granada PLoS One 2013, A.~Erzberger Proc Royal Soc 2013.
-
-\end{reviewer}
-
-The importance of amplitude in entrainment is an important consideration and has certainly been addressed by many studies in the past.
-We have incorporated several of these references into the motivation for the manuscript, as designing therapies to boost circadian amplitudes will surely have an effect on the entrainment of the oscillators as well:
-
-\begin{manuscript}{Page 2}
-The amplitude of circadian transcription is a relevant factor, and has been shown to play a critical role in phase resetting and entrainment (3, 4).
-Recent studies have further highlighted the importance of high peripheral clock amplitudes in maintaining metabolic health.
-\end{manuscript}
-
-\begin{reviewer}
-Ute Abraham et al. (Molecular Systems Biology) compare also coupled cells (SCN) with less coupled cells (lung) regarding their entrainment properties and discuss amplitude and relaxation effects.
-\end{reviewer}
-
-We have incorporated this reference into two places in the manuscript.
-First, in the introduction as reference (4) noted above, and also in the conclusion where we discuss the relative entrainment speed of coupled vs.
-uncoupled cells:
-
-\begin{manuscript}{Page 18}
-It is therefore interesting that liver cells have not developed a stronger mechanism of intercellular coupling, such as in the SCN, to maintain robust amplitudes in the absence
-of external cues.
-Perhaps the ability to quickly re-entrain to a phase shift, which is often slowed by coupling (4), has historically been more advantageous than protection against highly variable food intake.
+\begin{manuscript}{Page 7}
+To define an amplitude change metric for such a case, we compare a perturbed trajectory $x(t)$ to a phase-shifted limit cycle reference $y(t)$, for which $x(t) \to y(t)$ for sufficiently long times.
+Since $x(t)$ approaches the reference as $t \to \infty$, {\bfseries the means of both trajectories are equal and can be calculated by
+\begin{equation}
+  \mu \coloneqq \int_0^{2\pi} \frac{x^\gamma(\theta)}{2\pi} \; d\theta.
+  \tag{20}
+\end{equation}
+The amplitude change metric is defined as}
+\begin{equation}
+  \begin{aligned}
+    \Delta A (x(t), y(t)) &\coloneqq \int_0^\infty (x(t) - \mu)^2 - (y(t) - \mu)^2 \; dt\\
+    &= \int_0^\infty h(t) \; dt.
+  \end{aligned}
+  \tag{21}
+\end{equation}
 \end{manuscript}

### 2014-09-25 main_text.tex
starting second revision, adding changes requested by first reviewer

@@
 The choice of modeling framework dictates the type of amplitude response which can be predicted.
-Ordinary differential equation (ODE) models of gene regulation are capable of describing the amplitude and phase-resetting behavior of single cells, but fail to capture the collective dynamics of a population oscillators.
+Ordinary differential equation (ODE) models of gene regulation are capable of describing the amplitude and phase-resetting behavior of single cells, but fail to capture the collective dynamics of a population of oscillators.
 Likewise, phase-only models correctly capture the change in synchrony of a population, but do not capture fluctuations in amplitude of individual oscillators.
@@
 \begin{equation}
-  \frac{dx}{dt} = f(x(t), p)
+  \frac{dx}{dt} = f(x(t), p),
   \label{eq:odefn}
@@
 in which $x(t)$ represents the concentrations of the state variables, such as mRNA and protein concentrations, $f$ contains information on the production, degradation, and reactivity of the states, and $p$ are the kinetic parameters which govern reaction kinetics.
-Limit cycle models are ODE models in which the solution approaches a steady state oscillatory trajectory, satisfying the equation: \begin{equation}
-  \lim_{t \to \infty} \left[ x(t + T) - x(t) \right] = 0
+Limit cycle models are ODE models in which the solution approaches a steady state oscillatory trajectory, satisfying the equation:
+\begin{equation}
+  \lim_{t \to \infty} \left[ x(t + T) - x(t) \right] = 0.
   \label{eq:limit}
@@
 \begin{equation}
-  \tilde{t} = \frac{2\pi}{T}t; \quad \tilde{f} = \frac{T}{2\pi}\;f; \quad \frac{dx}{d\tilde{t}} = \tilde{f}(x(\tilde{t}), p)
+  \tilde{t} = \frac{2\pi}{T}t; \quad \tilde{f} = \frac{T}{2\pi}\;f; \quad \frac{dx}{d\tilde{t}} = \tilde{f}(x(\tilde{t}), p).
   \label{eq:that}
@@
 \begin{equation}
-  x(0) \coloneqq x^\gamma(\theta_0) + \Delta x(0)
+  x(0) \coloneqq x^\gamma(\theta_0) + \Delta x(0).
   \label{eq:stateperturbation}
@@
 \begin{equation}
-  \Delta x(\tilde{t}) = x(\tilde{t}) - x^\gamma(\tilde{t} + \theta_0)
+  \Delta x(\tilde{t}) = x(\tilde{t}) - x^\gamma(\tilde{t} + \theta_0).
   \label{eq:delxt}
@@
   x_{\tilde{d}}(-\tilde{d}) &= x^\gamma(\theta_0 - \tilde{d}) \\
-  \frac{dx_{\tilde{d}}}{d\tilde{t}} &= \tilde{f}(x_{\tilde{d}}(\tilde{t}), p + \Delta p) \label{eq:ode_pert}
+  \frac{dx_{\tilde{d}}}{d\tilde{t}} &= \tilde{f}(x_{\tilde{d}}(\tilde{t}), p + \Delta p) \label{eq:ode_pert}.
 \end{align}
@@
 \begin{equation}
-  \Delta x(0) = x_{\tilde{d}}(0) - x^\gamma(\theta_0)
+  \Delta x(0) = x_{\tilde{d}}(0) - x^\gamma(\theta_0).
 \end{equation}
@@
 The response of the system to very small perturbations is more efficiently calculated using ODE sensitivity analysis \cite{Rabitz1983}.
-The sensitivity matrix:
+The sensitivity matrix,
 \begin{equation}
-  S(t) = \frac{dx(t)}{dx(0)} = \lim_{\Delta x(0) \to 0}\frac{\Delta x(t)}{\Delta x(0)}
+  S(t) = \frac{dx(t)}{dx(0)} = \lim_{\Delta x(0) \to 0}\frac{\Delta x(t)}{\Delta x(0)},
   \label{eq:senslimit}
 \end{equation}
-can be calculated directly by integrating:
+can be calculated directly by integrating
 \begin{equation}
@@
 \begin{equation}
-  \frac{d}{d\tilde{t}}\frac{d\theta}{dp} = \frac{d\theta}{dx}\frac{d\tilde{f}}{dp} 
+  \frac{d}{d\tilde{t}}\frac{d\theta}{dp} = \frac{d\theta}{dx}\frac{d\tilde{f}}{dp}.
   \label{eq:pPRCequiv}
@@
 The usefulness of probability density functions in describing the phase and amplitude responses of populations of circadian cells has been previously shown \cite{Ukai2007}.
-As with all probability density functions:
+As with all probability density functions,
 \begin{equation}
-  \int_0^{2\pi} p(\theta, \tilde{t}) \; d\theta = 1
+  \int_0^{2\pi} p(\theta, \tilde{t}) \; d\theta = 1.
 \end{equation}
@@
 \begin{equation}
-  \frac{\partial p}{\partial \tilde{t}} = \frac{\partial p}{\partial \theta} + d\frac{\partial^2 p}{\partial \theta^2}\\
+  \frac{\partial p}{\partial \tilde{t}} = \frac{\partial p}{\partial \theta} + d\frac{\partial^2 p}{\partial \theta^2}.
   \label{eq:pde}
@@
   \frac{\partial p}{\partial \theta}(0, \tilde{t}) &= \frac{\partial p}{\partial \theta}(2\pi, \tilde{t}) \\
-  \text{IC:}\quad p(\theta, 0) &= \phi(\theta)\label{eq:pde_ic}
+  \text{IC:}\quad p(\theta, 0) &= \phi(\theta).
+  \label{eq:pde_ic}
 \end{align}
@@
   p(\theta, \tilde{t}) = \phi(\theta) * \mathcal{WN}(\theta; \tilde{t},
-  \sqrt{2\, d\, \tilde{t}})
+  \sqrt{2\, d\, \tilde{t}}),
 \end{equation}
-in which the wrapped normal distribution \cite{Mardia2009} is defined as:
+in which the wrapped normal distribution \cite{Mardia2009} is defined as
 \begin{equation}
@@
   \frac{1}{\sigma\sqrt{2\pi}}\sum_{k=-\infty}^\infty \exp\left[\frac{-(\theta
-  - \mu + 2\pi k)^2}{2\sigma^2}\right]
+  - \mu + 2\pi k)^2}{2\sigma^2}\right].
   \label{eq:wrapped_normal}
 \end{equation}
-Since the convolution of two normal distributions is also a normal distribution, it is efficient when possible to describe $\phi(\theta)$ as a normal distribution with mean $\mu_0$ and standard deviation $\sigma_0$, such that $p(\theta, \tilde{t})$ can be found analytically through:
+Since the convolution of two normal distributions is also a normal distribution, it is efficient when possible to describe $\phi(\theta)$ as a normal distribution with mean $\mu_0$ and standard deviation $\sigma_0$, such that $p(\theta, \tilde{t})$ can be found analytically through
 \begin{equation}
   p(\theta, \tilde{t}) = \mathcal{WN}(\theta; \mu_0 + \tilde{t},
-  \sqrt{\sigma_0^2 + 2d^2\tilde{t}^2})
+  \sqrt{\sigma_0^2 + 2d^2\tilde{t}^2}).
 \end{equation}
@@
 While such an amplitude change will only have a finite duration, it plays an important role when perturbations are repeatedly received by the clock, such as a peripheral oscillator entrained to daily metabolic stimuli.
-To define an amplitude change metric for such a case, we compare a perturbed trajectory $x(t)$ to a phase-shifted limit cycle reference $y(t)$, for which $x(t) \to y(t)$ for sufficiently long times:
+To define an amplitude change metric for such a case, we compare a perturbed trajectory $x(t)$ to a phase-shifted limit cycle reference $y(t)$, for which $x(t) \to y(t)$ for sufficiently long times.
+Since $x(t)$ approaches the reference as $t \to \infty$, the means of both trajectories are equal and can be calculated by
+\begin{equation}
+  \mu \coloneqq \int_0^{2\pi} \frac{x^\gamma(\theta)}{2\pi} \; d\theta.
+  \label{eq:mu}
+\end{equation}
+The amplitude change metric is defined as
 \begin{equation}
@@
     \Delta A (x(t), y(t)) &\coloneqq \int_0^\infty (x(t) - \mu)^2 - (y(t) - \mu)^2 \; dt\\
-    &= \int_0^\infty h(t) \; dt
+    &= \int_0^\infty h(t) \; dt.
   \end{aligned}
@@
 This amplitude metric was chosen over alternatives, such as peak-trough distance, for its analytical tractability.
-Since $x(t)$ approaches the reference as $t \to \infty$, the means of both trajectories are equal.
-The mean of the limit cycle can therefore be calculated as:
-\begin{equation}
-  \mu \coloneqq \int_0^{2\pi} \frac{x^\gamma(\theta)}{2\pi} \; d\theta
-  \label{eq:mu}
-\end{equation}
 The integrand in Eq.~\ref{eq:ampchangedefinition}, abbreviated by $h(t)$, compares the variances of the reference and perturbed trajectories.
@@
 Infinitesimal versions of response curves are often more general and easier to compute than those that track a specific perturbation strength \cite{Rand2004}.
-We therefore derive an expression for the infinitesimal ARC, defined as:
+We therefore derive an expression for the infinitesimal ARC, defined as
 \begin{equation}
-  \frac{dA}{dx} \coloneqq \lim_{\Delta x(0) \to 0} \frac{\Delta A\left(x(\tilde{t}),\ x^\gamma(\tilde{t} + \theta_0 + \Delta\theta)\right)}{\Delta x(0)} = \int_0^\infty \lim_{\Delta x(0) \to 0} \frac{h(t)}{\Delta x(0)} \; dt
+  \frac{dA}{dx} \coloneqq \lim_{\Delta x(0) \to 0} \frac{\Delta A\left(x(\tilde{t}),\ x^\gamma(\tilde{t} + \theta_0 + \Delta\theta)\right)}{\Delta x(0)} = \int_0^\infty \lim_{\Delta x(0) \to 0} \frac{h(t)}{\Delta x(0)} \; dt.
   \label{eq:sARC}
@@
   &= x^\gamma(t_\theta) + \tilde{f}\left(x^\gamma(t_\theta)\right)\Delta\theta +
-  O(\Delta\theta^2)
+  O(\Delta\theta^2).
 \end{align}
-Simplifying the integrand in Eq.~\ref{eq:sARC}:
+Simplifying the integrand in Eq.~\ref{eq:sARC},
 \begin{align}
@@
     x(\tilde{t}) + \tilde{f}\left(x^\gamma(t_\theta)\right)\Delta\theta +
-    2(x^\gamma(t_\theta) - \mu)\right) \right]
+    2(x^\gamma(t_\theta) - \mu)\right) \right].
 \end{align}
@@
   \lim_{\Delta x(0) \to 0} \frac{h(t)}{\Delta x(0)} &= 2\left(\lim_{\Delta x(0) \to 0}\frac{\Delta x(\tilde{t})}{\Delta x(0)} - \tilde{f}\left(x^\gamma(t_\theta)\right)\lim_{\Delta x(0) \to 0}\frac{\Delta\theta}{\Delta x(0)}\right) \left(x^\gamma(t_\theta) - \mu\right)\\
-  &= 2\left(S(\tilde{t}) - \tilde{f}(x^\gamma(t_\theta))\frac{d\theta}{dx}\right)\left(x^\gamma(t_\theta) - \mu\right)
+  &= 2\left(S(\tilde{t}) - \tilde{f}(x^\gamma(t_\theta))\frac{d\theta}{dx}\right)\left(x^\gamma(t_\theta) - \mu\right).
   \label{eq:sens_sub}
@@
 \begin{equation}
-  \frac{dA}{dx} = \int_0^\infty 2\left(S(\tilde{t}) - \tilde{f}(x^\gamma(t_\theta))\frac{d\theta}{dx}\right)\left(x^\gamma(t_\theta) - \mu\right) \; dt
+  \frac{dA}{dx} = \int_0^\infty 2\left(S(\tilde{t}) - \tilde{f}(x^\gamma(t_\theta))\frac{d\theta}{dx}\right)\left(x^\gamma(t_\theta) - \mu\right) \; dt.
   \label{eq:siARC}
@@
 The second term $(x^\gamma - \mu)$ weights this distance by whether or not the deviation occurs above or below the oscillatory mean, yielding negative amplitude changes when the trajectory is perturbed closer to the mean.
-Just as with the parameter-impulse PRC, the infinitesimal parameter-impulse ARC is defined as:
+Just as with the parameter-impulse PRC, the infinitesimal parameter-impulse ARC is defined as
 \begin{equation}
-  \frac{d}{d\tilde{t}}\frac{dA}{dp} \coloneqq \lim_{\tilde{d},\; \Delta p \to 0} \frac{\Delta A}{\tilde{d}\, \Delta p}
+  \frac{d}{d\tilde{t}}\frac{dA}{dp} \coloneqq \lim_{\tilde{d},\; \Delta p \to 0} \frac{\Delta A}{\tilde{d}\, \Delta p},
 \end{equation}
@@
 \begin{equation}
-  \frac{d}{d\tilde{t}}\frac{dA}{dp} = \frac{dA}{dx}\frac{d\tilde{f}}{dp}
+  \frac{d}{d\tilde{t}}\frac{dA}{dp} = \frac{dA}{dx}\frac{d\tilde{f}}{dp}.
     \label{eq:piARC}
@@
 \begin{equation}
-  \hat{p}(\theta, \tilde{t})\; dg(\theta) = p(\theta, \tilde{t})\; d\theta
+  \hat{p}(\theta, \tilde{t})\; dg(\theta) = p(\theta, \tilde{t})\; d\theta.
   \label{eq:pdfinversion}
@@
 For $\rho = 1$, the population is clustered about one mean phase, while for $\rho = 0$ the population is evenly balanced across the unit circle.
-Complex variables for the population before and after perturbation can be calculated via:
+Complex variables for the population before and after perturbation can be calculated via
 \begin{align}
@@
   \hat{z} &\coloneqq  \int_0^{2\pi} e^{i\theta} \hat{p}(\theta, \tilde{t}) \; d\theta =
-  \int_0^{2\pi} e^{i g(\theta)} p(\theta, \tilde{t}) \; d\theta
+  \int_0^{2\pi} e^{i g(\theta)} p(\theta, \tilde{t}) \; d\theta.
   \label{eq:zhat}
@@
 In Eq.~\ref{eq:zhat}, we avoid calculating the perturbed population explicitly by instead integrating over the new phases at the prior population density function.
-Population-level amplitude and phase responses can therefore be calculated by:
+Population-level amplitude and phase responses can therefore be calculated by
 \begin{align}
   \Delta \bar{\theta} &= \angle z - \angle \hat{z} \\
-  \Delta \rho &= |z| - |\hat{z}|
+  \Delta \rho &= |z| - |\hat{z}|.
   \label{eq:popampchange}
@@
 \begin{equation}
-  \bar{x}(\tilde{t}) = \int_0^{2\pi} x^\gamma(\theta) p(\theta, \tilde{t}) \;
-  d\theta
+  \bar{x}(\tilde{t}) = \int_0^{2\pi} x^\gamma(\theta) p(\theta, \tilde{t}) \; d\theta.
   \label{eq:xbar}
@@
 \begin{align}
-  \bar{x}(\tilde{t}) &= \frac{1}{\sqrt{4\pi d\tilde{t}}}\int_{-\infty}^{\infty}
-  \cos(x) \exp\left(-\frac{(x - \tilde{t})^2}{4d\tilde{t}}\right)\; dx \\
-  &= \Re\left[\frac{1}{\sqrt{4\pi d\tilde{t}}}\int_{-\infty}^{\infty}
-    \exp(ix) \exp\left(-\frac{(x - \tilde{t})^2}{4d\tilde{t}}\right)\; dx\right] \\
+  \bar{x}(\tilde{t}) &= \frac{1}{\sqrt{4\pi d\tilde{t}}}\int_{-\infty}^{\infty} \cos(x) \exp\left(-\frac{(x - \tilde{t})^2}{4d\tilde{t}}\right)\; dx \\
+  &= \Re\left[\frac{1}{\sqrt{4\pi d\tilde{t}}}\int_{-\infty}^{\infty} \exp(ix) \exp\left(-\frac{(x - \tilde{t})^2}{4d\tilde{t}}\right)\; dx\right] \\
     &= \Re\left[e^{(i - d)\tilde{t}}\right]\\
-    &= e^{-d\tilde{t}} \cos(\tilde{t}) \label{eq:expsin}
+    &= e^{-d\tilde{t}} \cos(\tilde{t}).
+    \label{eq:expsin}.
 \end{align}
@@
 First, for long times after the perturbation, each individual oscillator will have returned to the limit cycle $x^\gamma(\theta)$, but with a new population density $\hat{p}(\theta, \tilde{t})$.
-This steady-state perturbed trajectory $\hat{x}_{ss}(\tilde{t})$ can be found by:
+This steady-state perturbed trajectory $\hat{x}_{ss}(\tilde{t})$ can be found by
 \begin{equation}
-  \hat{x}_{ss}(\tilde{t}) = \int_0^{2\pi} x^\gamma(\theta) \hat{p}(\theta, \tilde{t}) \; d\theta
+  \hat{x}_{ss}(\tilde{t}) = \int_0^{2\pi} x^\gamma(\theta) \hat{p}(\theta, \tilde{t}) \; d\theta.
   \label{eq:xhatss}
@@
   \delta x(\theta_0, \tilde{t}) &\coloneqq x(\tilde{t}) - x^\gamma(\tilde{t} + \theta_0 + \Delta \theta) \\
-  \therefore\; \lim_{\tilde{t} \to \infty} \delta x(\theta, \tilde{t}) &= 0
+  \therefore\; \lim_{\tilde{t} \to \infty} \delta x(\theta, \tilde{t}) &= 0.
 \end{align}
@@
 It is therefore more straightforward to find the average effect of single-cell perturbations at the population level by weighting the deviations by the phase density function prior to perturbation.
-The population-level response to a perturbation, $\hat{x}(\tilde{t})$, is therefore defined as:
+The population-level response to a perturbation, $\hat{x}(\tilde{t})$, is therefore defined as
 \begin{equation}
-  \hat{x}(\tilde{t}) = \int_0^{2\pi} x^\gamma(\theta)\hat{p}(\theta, \tilde{t}) + \delta x(\theta, \tilde{t})p(\theta, \tilde{t}) \; d\theta
+  \hat{x}(\tilde{t}) = \int_0^{2\pi} x^\gamma(\theta)\hat{p}(\theta, \tilde{t}) + \delta x(\theta, \tilde{t})p(\theta, \tilde{t}) \; d\theta.
   \label{eq:xhat}

### 2014-09-26 response_letter.tex
Addressing reviewer 2's last point, adding fourth supplemental figure

@@
 
+\section*{Reviewer \#2}
+\begin{reviewer}
+The revised manuscript becomes much more clear. The authors addressed most of my concerns in their revision. However, I have still couple concerns regarding the manuscript.
+
+1) In the previous review, I was concerned about using the Gillesplie algorithm for the model with non-elementary reactions because recent studies have clearly showed the inaccuracy of doing that (Thomas, Straube and Grima BMC Syst Biol (2012), Agarwal et al.
+J Chem Phy (2012) and Kim, Josic and Bennett Biophy J (2014)).
+Regarding this concern, the author showed that their continuous approximation also works for the model based on mass-action kinetics (Model 2).
+But, in the manuscript, the approximation is more accurate for Model 2 than other models, which are based on non-elementary reactions.
+This indicates that the inaccuracy coming from the Gillespie algorithm with non-elementary reaction appears to contribute the error seen in Model 1 and 3.
+Furthermore, I am mostly concerned that this manuscript can mislead the audience that using Gillespie algorithm with non-elementary reactions is justified.
+Please, clearly discuss the limitation in this approach.
+\end{reviewer}
+
+\begin{reviewer}
+2) Another concern is regarding the major result of the manuscript (Fig.
+5).
+The author perturbed the degradation of Per mRNA parameter in response to light.
+However, this conflicts with the previous experimental studies, which have shown that light promotes the transcription of Per mRNA via CREB, not degradation.
+\end{reviewer}
+
+\begin{reviewer}
+3) The key goal of this manuscript is distinguishing the single-cell level and population-level amplitude change.
+For this, the authors applied their tool to the experimental data (Fig.
+5).
+However, Fig.
+5 does not clearly show how their tool achieves such goal.
+Please provide additional figures.
+For instance, $x_{ss}$ (steady-state perturbed trajectory) for perturbed and unperturbed cases, $p(\theta, t)$ for unperturbed case as well as perturbed case etc, so that readers can clearly see the difference of effect via single cell level and population level.
+\end{reviewer}
+
+This is a good suggestion, and we have therefore added a figure to explicitly highlight the contributions from the population and single-cell level.
+Due to lack of space in the main text and desire not to further complicate the original figure, we have added this figure as Figure~S4.
+A reference to the figure was added to the main text of the manuscript:
+
+\begin{manuscript}{Page 16}
+The phase probability density function for the light-sensitive model trajectory is shown in Fig.~5C for several representative time points.
+\ldots
+{\bfseries Amplitude change in this case is mediated both at the population and single-cell level, as indicated by their representative ARCs.
+The contributions from each source are summarized in Fig.~S4.}
+\end{manuscript}
+
+with the following figure and caption added to the supplementary material:
+
+\begin{manuscript}{Supplemental Info, Page 7}
+  \begin{center}
+    \includegraphics[width=.85\textwidth]{figures/figure_S4.pdf}\\
+  \end{center}
+{\bfseries Figure S4: Evidence of single-cell and population-level amplitude change.}
+({\bfseries A}) The simulated control trajectory as shown in Fig.~5 is plotted again as $\bar{x}(\tilde{t})$.
+The simulated light-sensitive trajector is also replotted as $\hat{x}(\tilde{t})$, which incorperates both single-cell and population-level effects.
+The difference between this trajectory and the steady-state perturbed trajectory following both pulses, $\hat{x}_{ss}(\tilde{t})$, demonstrates the effect of single-cell level amplitude change in this particular case.
+Single-cell amplitudes are greatly increased following the first pulse, and altered significantly following the second.
+({\bfseries B})
+Comparison of the population densities for the control (dashed) and perturbed (solid) simulations.
+The perturbed population is greatly desynchronized following the first light pulse, as indicated by its flatness relative to the control trajectory. 
+Following the second pulse, the perturbed population is more concentrated than the control, indicating higher amplitudes. 
+These comparisons demonstrate the changes in population synchrony which occur due to the light pulse. 
+\end{manuscript}
+
 \end{document}

### 2014-09-26 supplemental_info.tex
Addressing reviewer 2's last point, adding fourth supplemental figure

@@
 
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=.75\textwidth]{figures/figure_S4.pdf}
+    \caption{
+{\bfseries Evidence of single-cell and population-level amplitude change.}
+({\bfseries A}) The simulated control trajectory as shown in Fig.~5 is plotted again as $\bar{x}(\tilde{t})$.
+The simulated light-sensitive trajector is also replotted as $\hat{x}(\tilde{t})$, which incorperates both single-cell and population-level effects.
+The difference between this trajectory and the steady-state perturbed trajectory following both pulses, $\hat{x}_{ss}(\tilde{t})$, demonstrates the effect of single-cell level amplitude change in this particular case.
+Single-cell amplitudes are greatly increased following the first pulse, and altered significantly following the second.
+({\bfseries B})
+Comparison of the population densities for the control (dashed) and perturbed (solid) simulations.
+The perturbed population is greatly desynchronized following the first light pulse, as indicated by its flatness relative to the control trajectory. 
+Following the second pulse, the perturbed population is more concentrated than the control, indicating higher amplitudes. 
+These comparisons demonstrate the changes in population synchrony which occur due to the light pulse. 
+}
+\end{center}
+\end{figure}
+
 \clearpage

### 2014-09-26 response_letter.tex
Rebutting second point

@@
 
+This is a valid point, since our analysis captures the increase in {\itshape Per} mRNA following light pulses as a {\itshape\bfseries decrease} in the degradation rate of {\itshape Per}.
+In the model, {\itshape Per} transcription is controlled by several parameters: the maximum transcription rate, transcription rate kinetic constant, maximum degradation rate, degradation rate kinetic constant, and translation rate.
+Since we acknowledge that the model does not perfectly describe the true system, we tried to fit the model to data by searching the parameter ARCs for parameters and reporter states which gave the appropriate response.
+From these parameters, the maximum {\itshape Per} transcription rate and {\itshape Cry} mRNA emerged as the best choice given the response of the model and known experimental results:
+
+\begin{manuscript}{Page 16}
+The next step in capturing the experimental data was finding a perturbation capable of desynchronizing the system.
+To find such a perturbation, we calculated population-level ARCs at several different knockdown strengths for each parameter in the model.
+We selected the degradation rate of {\itshape Per} mRNA from the feasible parameters, due to PER's known induction by CREB following photo-perturbation (46).
+The parameter was reduced by $28.5\%$ during the light pulses, a value fit to maximize light-induced desynchrony.
+The output variable, PER2-luciferase in the experimental system, was chosen to be represented by {\itshape Cry1} mRNA - an E box activated gene that is buffered from the direct effects of the parameter perturbation.
+\end{manuscript}
+
+% We should emphasize that we are not predicting {\itshape Per} induction following light actually occurs via degradation, but rather that by acknowledging the model does not perfectly capture experimental
+
 \begin{reviewer}

### 2014-09-26 response_letter.tex
Addressing first concern, legitimacy of the stochastic simulation algorithm

@@
 
-We thank the reviewers for their second round of comments of the manuscript, in particular Reviewer \#2 for their continued attention to detail.
+We thank the reviewers for their second round of comments on the manuscript, in particular Reviewer \#2 for their continued attention to detail.
 We believe addressing these remaining concerns and suggestions have improved the polish of the manuscript.
@@
 
+We acknowledge there has been a great deal of literature concerning the accuracy of using the Gillespie stochastic simulation algorithm for non-elementary propensity functions.
+In K. R. Sanft, {\itshape et al, IET Syst. Biol.} (2011), the authors conclude that the Michealis-Menten approximation overestimates the noise of the corresponding full system of equations.
+In L.
+A.
+Widmer, {\itshape et al, J.
+Chem.
+Phys.} (2013), the authors prove that the noise of the full system is bounded between $\nicefrac{3}{4} \to 1$ of the noise estimated by the Michealis-Menten approximation.
+For hill-type kinetics, the full system can take an arbitrary level of noise relative to the approximated system (L. A. Widmer, Master's Thesis, 2013).
+
+However, the most important consideration here is that the full system of equations require additional parameter values to be chosen; by choosing these parameter values, the noise of the full system can be changed relative to the system with non-elementary propensity functions.
+Since the values of these expanded parameters are not specified in the models used, we simply fit the noise characteristics of the approximated model (using the volume parameter) to yield physiologically realistic desynchronization rates.
+Therefore, the validity of whether or not these approximated solutions closely represent the noise characteristics of the full solutions does not play a large role in determining the accuracy of the continuous approximation.
+
+To address the larger point, we discuss reasons why the continuous approximations are less accurate for the larger model than the smaller model. 
+We acknowledge that the method is only completely accurate in the limit of no noise in the underlying oscillator.
+When noise is present, the `unperturbed' population consists of oscillators not only with a distribution of phases, but also with the states of many oscillators far from the deterministic limit cycle (this can be seen in the left plot of supplemental movies S1 \& S2).
+As a result, the responses of oscillators not directly on the limit cycle to a pulse can differ from their limit cycle counterparts, decreasing the accuracy of the method for systems with high noise.
+In practice, however, it takes a considerable amount of noise before the continuous predictions deviate substantially from the stochastic populations.
+It is also important to consider that the larger model consists of many more state variables than the mass-action model, only one of which is shown in Fig.~S3.
+Therefore, the difference in accuracy likely comes from increased deviation about the deterministic limit cycle in 8 spatial dimensions, rather than from non-elementary propensity functions.
+
+We have endeavored to make these points more clear in the manuscript:
+
+\begin{manuscript}{Supplemental Info, Page 6}
+Good agreement is seen between the continuous and stochastic simulations, indicating the proposed method is suitable for predicting the population-level responses for a variety of model types.
+{\bfseries The slightly reduced accuracy seen in the larger Model 2 demonstrates a limitation of the method, which should be considered before applying the method to extreme cases.
+As Eq.~44 tabulates the response of the system from the limit cycle, systems which are perturbed from states far from the deterministic limit cycle will not be captured appropriately.
+Therefore, the difference in accuracy between Model 2 and Model 3 likely comes from increased deviation about the deterministic limit cycle in a greater number of spatial dimensions.}
+\end{manuscript}
+
+Additionally, we caution readers against the unchecked use of the Gillespie algorithm for systems with non-elementary propensity functions:
+
+\begin{manuscript}{Supplemental Info, Page 6}
+{\bfseries Model 1:}
+Adapted from Nov\'{a}k \& Tyson, 2008 (1). Model used for
+ARC demonstrations in Figs. 1-3, S1 ($P=4$) and Fig. S2, Movies S1 \& S2
+($P=2$).
+{\bfseries It should be noted that the use of the stochastic simulation algorithm for reactions with non-elementary propensities may inaccurately represent the noise of the full system.
+In this case, the volume of the stochastic simulations was chosen as $\Omega = 250$ to give a physiologically realistic desynchronization rate.}
+\end{manuscript}
+
 \begin{reviewer}
@@
 
-% We should emphasize that we are not predicting {\itshape Per} induction following light actually occurs via degradation, but rather that by acknowledging the model does not perfectly capture experimental
+We should emphasize that we are not predicting {\itshape Per} induction following light actually occurs via degradation, but rather that by acknowledging the model does not perfectly capture the experimental system, we simply seek a qualitatively accurate description of the underlying phenomena.
 
@@
 Comparison of the population densities for the control (dashed) and perturbed (solid) simulations.
-The perturbed population is greatly desynchronized following the first light pulse, as indicated by its flatness relative to the control trajectory. 
+The perturbed population is greatly desynchronized following the first light pulse, as indicated by its flatness relative to the control population. 
 Following the second pulse, the perturbed population is more concentrated than the control, indicating higher amplitudes. 

### 2014-09-26 supplemental_info.tex
Addressing first concern, legitimacy of the stochastic simulation algorithm

@@
 \begin{model}
-\caption{Adapted from Nov\'{a}k \& Tyson, 2008 \cite{Novak2008}. Model used for
+\caption{
+Adapted from Nov\'{a}k \& Tyson, 2008 \cite{Novak2008}. Model used for
 ARC demonstrations in Figs. 1-3, S1 ($P=4$) and Fig. S2, Movies S1 \& S2
-($P=2$).}
+($P=2$).
+It should be noted that the use of the stochastic simulation algorithm for reactions with non-elementary propensities may inaccurately represent the noise of the full system.
+In this case, the volume of the stochastic simulations was chosen as $\Omega = 250$ to give a physiologically realistic desynchronization rate.
+}
   \centering
@@
     \includegraphics[width=\textwidth]{figures/figure_S3.pdf}
-    \caption{{\bfseries Validation of continuous methods with alternate
-    models.} ({\bfseries A}) The Oregonator model (Model 3) was tested as an
-    example of a mass-action limit cycle oscillator. A population of 2000
-    oscillators was perturbed at size different mean phases, $\mu$, by
-    increasing the $\mathit{c2}$ parameter by $50\%$ for $d=\pi$ (highlighted
-    region). Only the average $\mathit{Y2}$ variable is plotted. ({\bfseries
-    B}) The model described in \cite{Hirota2012} (Model 2) was similarly tested
-    at a variety of mean phases by reducing the $\mathit{vdp}$ parameter by
-    $28.5\%$ and plotting the resulting $\mathit{c2}$ state variable (as in
-    Fig.~5). Parameters for the continuous approximations were found by
-    estimating the phase, initial standard deviation, period, and phase
-    diffusivity of the control population. Good agreement is seen between the
-  continuous and stochastic simulations, indicating the proposed method is
-suitable for predicting the population-level responses for a variety of model
-types.}
+    \caption{
+{\bfseries Validation of continuous methods with alternate models.}
+({\bfseries A}) The Oregonator model (Model 3) was tested as an example of a mass-action limit cycle oscillator.
+A population of 2000 oscillators was perturbed at size different mean phases, $\mu$, by increasing the $\mathit{c2}$ parameter by $50\%$ for $d=\pi$ (highlighted region).
+Only the average $\mathit{Y2}$ variable is plotted.
+({\bfseries B}) The model described in \cite{Hirota2012} (Model 2) was similarly tested at a variety of mean phases by reducing the $\mathit{vdp}$ parameter by $28.5\%$ and plotting the resulting $\mathit{c2}$ state variable (as in Fig.~5).
+Parameters for the continuous approximations were found by estimating the phase, initial standard deviation, period, and phase diffusivity of the control population.
+Good agreement is seen between the continuous and stochastic simulations, indicating the proposed method is suitable for predicting the population-level responses for a variety of model types.
+The slightly reduced accuracy seen in the larger Model 2 demonstrates a limitation of the method, which should be considered before applying the method to extreme cases.
+As Eq.~44 tabulates the response of the system from the limit cycle, systems which are perturbed from states far from the deterministic limit cycle will not be captured appropriately.
+Therefore, the difference in accuracy between Model 2 and Model 3 likely comes from increased deviation about the deterministic limit cycle in a greater number of spatial dimensions.
+}
 \end{center}

### 2014-09-27 response_letter.tex
Finished john's edits, sending to stephanie

@@
 We thank the reviewers for their second round of comments on the manuscript, in particular Reviewer \#2 for their continued attention to detail.
-We believe addressing these remaining concerns and suggestions have improved the polish of the manuscript.
+We believe addressing these remaining concerns and suggestions have greatly improved the manuscript.
 
@@
 
-This typo has been addressed, the sentence now reads:
+This typo has been addressed, and the sentence now reads:
 
@@
 
-We acknowledge there has been a great deal of literature concerning the accuracy of using the Gillespie stochastic simulation algorithm for non-elementary propensity functions.
+This is a fair criticism, as we did not intend to defend the use of the stochastic simulation algorithm for non-elementary propensity functions. 
+However, we believe the use of the algorithm in this case is justified, and is not the main source of error in the continuous approximation.
+We provide a more detailed defense of our use of the algorithm for the reviewer's specific concerns below, and have incorporated additional disclaimers in the text to warn readers against the use non-elementary propensity functions in the general case.
+
+There has indeed been a great deal of literature concerning the accuracy of using the Gillespie stochastic simulation algorithm for non-elementary propensity functions.
 In K. R. Sanft, {\itshape et al, IET Syst. Biol.} (2011), the authors conclude that the Michealis-Menten approximation overestimates the noise of the corresponding full system of equations.
@@
 Chem.
-Phys.} (2013), the authors prove that the noise of the full system is bounded between $\nicefrac{3}{4} \to 1$ of the noise estimated by the Michealis-Menten approximation.
-For hill-type kinetics, the full system can take an arbitrary level of noise relative to the approximated system (L. A. Widmer, Master's Thesis, 2013).
+Phys.} (2013), the authors prove that the noise of the full system is bounded on $[\nicefrac{3}{4}, 1]$ of the noise estimated by the Michealis-Menten approximation.
+For Hill-type kinetics, the full system can take an arbitrary level of noise relative to the approximated system (L. A. Widmer, Master's Thesis, 2013).
 
-However, the most important consideration here is that the full system of equations require additional parameter values to be chosen; by choosing these parameter values, the noise of the full system can be changed relative to the system with non-elementary propensity functions.
-Since the values of these expanded parameters are not specified in the models used, we simply fit the noise characteristics of the approximated model (using the volume parameter) to yield physiologically realistic desynchronization rates.
+The most important consideration is that converting the models to a full mass-action description of the system (or to a slightly less detailed total quasi-steady-state approximation as discussed in the literature cited by the reviewer) would require additional parameter values to be chosen.
+This parameter set is not identifiable, as an infinite number of parameter choices exist that will yield equivalent continuous dynamics to the Hill approximation.
+Each of these parameter sets will have different noise characteristics, which may not be reflected by the arbitrary noise characteristics arising from the Hill simplification.
+However, since we do not specify these expanded parameters, we simply fit the noise characteristics of the approximated model (using the volume parameter) to yield physiologically realistic desynchronization rates.
 Therefore, the validity of whether or not these approximated solutions closely represent the noise characteristics of the full solutions does not play a large role in determining the accuracy of the continuous approximation.
 
-To address the larger point, we discuss reasons why the continuous approximations are less accurate for the larger model than the smaller model. 
-We acknowledge that the method is only completely accurate in the limit of no noise in the underlying oscillator.
-When noise is present, the `unperturbed' population consists of oscillators not only with a distribution of phases, but also with the states of many oscillators far from the deterministic limit cycle (this can be seen in the left plot of supplemental movies S1 \& S2).
+To address the larger point, we discuss reasons why our continuous approximations are less accurate for the larger model than the smaller model. 
+We first point out that the method is an asymptotic approximation of a population of oscillators in the limit of no noise, as we account for the noise-induced desynchronization of oscillators but not noise-induced deviation from the limit cycle.
+When noise is present, the `unperturbed' population consists of oscillators not only with a distribution of phases, but also with the states of each oscillator in a region of space near the deterministic limit cycle (this can be seen in the left plot of supplemental movies S1 \& S2).
 As a result, the responses of oscillators not directly on the limit cycle to a pulse can differ from their limit cycle counterparts, decreasing the accuracy of the method for systems with high noise.
@@
 {\bfseries It should be noted that the use of the stochastic simulation algorithm for reactions with non-elementary propensities may inaccurately represent the noise of the full system.
-In this case, the volume of the stochastic simulations was chosen as $\Omega = 250$ to give a physiologically realistic desynchronization rate.}
+In this case, we simply fit the noise characteristics of the approximated model (using the volume parameter) to yield physiologically realistic desynchronization rates.
+The volume of the stochastic simulations was chosen as $\Omega = 250$.}
 \end{manuscript}
@@
 Since we acknowledge that the model does not perfectly describe the true system, we tried to fit the model to data by searching the parameter ARCs for parameters and reporter states which gave the appropriate response.
-From these parameters, the maximum {\itshape Per} transcription rate and {\itshape Cry} mRNA emerged as the best choice given the response of the model and known experimental results:
+From these parameters, the maximum {\itshape Per} transcription rate and {\itshape Cry} mRNA emerged as the best choice given the response of the model and known experimental results.
+We have added the following text to the manuscript to make this point more apparent:
 
@@
 To find such a perturbation, we calculated population-level ARCs at several different knockdown strengths for each parameter in the model.
-We selected the degradation rate of {\itshape Per} mRNA from the feasible parameters, due to PER's known induction by CREB following photo-perturbation (46).
+We selected the degradation rate of {\itshape Per} mRNA from the feasible parameters due to PER's known induction by CREB following photo-perturbation (46). 
+{\bfseries While not an exact match of the experimental system, reducing degradation rate yields a similar response to increasing transcription.}
 The parameter was reduced by $28.5\%$ during the light pulses, a value fit to maximize light-induced desynchrony.
@@
 
-This is a good suggestion, and we have therefore added a figure to explicitly highlight the contributions from the population and single-cell level.
+This is a good suggestion, and we thank the reviewer for helping us clarify our results.
+We have added a figure to explicitly highlight the contributions from the population and single-cell level.
 Due to lack of space in the main text and desire not to further complicate the original figure, we have added this figure as Figure~S4.
@@
 
+We would like to once again thank all our reviewers, as we feel our manuscript has benefited greatly from their suggestions and critiques.
+
 \end{document}

### 2014-09-27 main_text.tex
Addressing reviewer 2's last point, adding fourth supplemental figure
Finished john's edits, sending to stephanie

@@
 To find such a perturbation, we calculated population-level ARCs at several different knockdown strengths for each parameter in the model.
-We selected the degradation rate of {\itshape Per} mRNA from the feasible parameters, due to PER's known induction by CREB following photo-perturbation \cite{Tischkau2003}.
+We selected the degradation rate of {\itshape Per} mRNA from the feasible parameters due to PER's known induction by CREB following photo-perturbation \cite{Tischkau2003}.
+While not an exact match of the experimental system, reducing degradation rate yields a similar response to increasing transcription.
 The parameter was reduced by $28.5\%$ during the light pulses, a value fit to maximize light-induced desynchrony.
@@
 As higher-frequency features, these bimodal distributions dissipate as the phases diffuse.
+Amplitude change in this case is mediated both at the population and single-cell level, as indicated by their representative ARCs. The contributions from each source are summarized in Fig.~S4.
 

### 2014-09-27 supplemental_info.tex
Finished john's edits, sending to stephanie

@@
 It should be noted that the use of the stochastic simulation algorithm for reactions with non-elementary propensities may inaccurately represent the noise of the full system.
-In this case, the volume of the stochastic simulations was chosen as $\Omega = 250$ to give a physiologically realistic desynchronization rate.
+In this case, we simply fit the noise characteristics of the approximated model (using the volume parameter) to yield physiologically realistic desynchronization rates.
+The volume of the stochastic simulations was chosen as $\Omega = 250$.
 }

### 2014-09-29 response_letter.tex
Adding stephanie's change

@@
 When noise is present, the `unperturbed' population consists of oscillators not only with a distribution of phases, but also with the states of each oscillator in a region of space near the deterministic limit cycle (this can be seen in the left plot of supplemental movies S1 \& S2).
-As a result, the responses of oscillators not directly on the limit cycle to a pulse can differ from their limit cycle counterparts, decreasing the accuracy of the method for systems with high noise.
+As a result, the response to a pulse of an oscillator off the limit cycle can differ from the response of an on-limit oscillator with equivalent phase, decreasing the accuracy of the method for systems with high noise.
 In practice, however, it takes a considerable amount of noise before the continuous predictions deviate substantially from the stochastic populations.

### 2014-11-06 supplemental_info.tex
reflowing text in supplement
proofs recieved, some changes

@@
 
+% Puts the affiliations on the cover page
+\usepackage{authblk}
+
 \begin{document}
+\title{Supporting Material for:\\ Amplitude metrics for cellular \\ circadian bioluminescence reporters}
+\author[1]{Peter C. St. John}
+\author[2]{Stephanie R. Taylor}
+\author[1]{John H. Abel}
+\author[1,*]{Francis J. Doyle III}
+\affil[1]{Department of Chemical Engineering, University of California Santa
+Barbara, Santa Barbara, California 93106-5080}
+\affil[2]{Department of Computer Science, Colby College, Waterville, Maine
+04901}
+\affil[*]{Email: \texttt{doyle@engineering.ucsb.edu}}
+\date{}
+\maketitle
 
@@
 
-\tablehead{\toprule Parameter & Value\\\midrule}
-
-\begin{supertabular}{rl}
+% \tablehead{}
+\begin{tabular}{rl}
+\toprule Parameter & Value\\\midrule
 $vtp $  & 0.195   \\
@@
 $vdp $  & 0.326   \\
-$vdc1$  & 0.676   \\
-\bottomrule\end{supertabular}\hfil
-\begin{supertabular}{rl}
+$vdc1$  & 0.676   \\\bottomrule
+\end{tabular}\hfil
+%
+\begin{tabular}{rl}
+\toprule Parameter & Value\\\midrule
 $vdc2$ & 0.608   \\
@@
 $vdC2$ & 1.52    \\
-$vdCn$ & 1.69    \\
-\bottomrule\end{supertabular}\hfil
-\begin{supertabular}{rl}
+$vdCn$ & 1.69    \\\bottomrule
+\end{tabular}\hfil
+%
+\begin{tabular}{rl}
+\toprule Parameter & Value\\\midrule
 $MC2n $ & 2.02    \\
@@
 $vdC1P$ & 0.00175 \\
-$ktxnp$ & 3.0     \\
-\bottomrule\end{supertabular}
+$ktxnp$ & 3.0     \\\bottomrule
+\end{tabular}
 \end{model}
@@
 \clearpage
-{\bfseries Movie S1:} Stochastic simulation of a two-state limit
-cycle oscillator. A population of 225 stochastic oscillators was simulated
-using the Gillespie stochastic simulation algorithm (SSA). The deterministic
-limit cycle, $x^\gamma(\hat{t})$, is shown in black (left), and the current
-location of individual stochastic oscillators are shown by green dots. The
-population average, plotted in both the state (left) and time (right) domains,
-is shown in red. Stochastic noise drives the population to gradually
-desynchronize, resulting in damped oscillations of the average protein level.
+{\bfseries Movie S1:} Stochastic simulation of a two-state limit cycle oscillator.
+A population of 225 stochastic oscillators was simulated using the Gillespie stochastic simulation algorithm (SSA).
+The deterministic limit cycle, $x^\gamma(\hat{t})$, is shown in black (left), and the current location of individual stochastic oscillators are shown by green dots.
+The population average, plotted in both the state (left) and time (right) domains, is shown in red.
+Stochastic noise drives the population to gradually desynchronize, resulting in damped oscillations of the average protein level.
 \\[2ex]
 
-{\bfseries Movie S2:} Perturbation to a population of stochastic
-oscillators. The population of oscillators is perturbed by a $50\%$ reduction
-in the protein translation rate to each of the oscillators from $\hat{t} =
--\nicefrac{\pi}{4} \to 0$ (yellow frames), resulting in both single-cell and
-population-level amplitude change. The oscillators quickly recover to the limit
-cycle (single-cell level amplitude change), but the population is left
-desynchronized (population-level amplitude change).
+{\bfseries Movie S2:} Perturbation to a population of stochastic oscillators.
+The population of oscillators is perturbed by a $50\%$ reduction in the protein translation rate to each of the oscillators from $\hat{t} = -\nicefrac{\pi}{4} \to 0$ (yellow frames), resulting in both single-cell and population-level amplitude change.
+The oscillators quickly recover to the limit cycle (single-cell level amplitude change), but the population is left desynchronized (population-level amplitude change).
 
